---
title: "Comparing genomic offset predictions"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 11px
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
knitr::opts_chunk$set(cache=F, size="tiny")
options(width = 300)
library(knitr)
library(kableExtra)
library(tidyverse)
library(here)
library(cowplot)
library(magrittr)
library(corrplot)
library(RColorBrewer)

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
```

We compare the genomic offset predictions from the four following methods:

  - Generalized dissimilarity modelling (GDM). See report `9_GeneralizedDissimilarityModelling_GenomicOffsetPredictions.html`.
  
  - Latent factor mixed model (LFMM) . See report `7_LFMM_IdentificationCandidateSNPs_GenomicOffsetPredictions.html`.
  
  - Gradient Forest (GF). See report `10_GradientForest_GenomicOffsetPredictions.html`.
  
  - Redundancy analysis (RDA). See report `11_RedundancyAnalysis_GenomicOffsetPredictions.html`.

```{r}
# Method names
meth_names <- c("GDM","LFMM","GF","RDA", "pRDA")

# population names
pop_names <- readRDS(here("data/GenomicData/MainGenePoolPopulations.rds")) %>%  arrange(pop) %>% pull(pop)

# SNP sets
snp_sets <- readRDS(here("outputs/list_snp_sets.rds"))
```

# Comparing population ranks

In this section, we compare GO predictions from the different methods (GF, GDM, RDA and LFMM), SNP sets (candidates, control and all SNPs) and GCMs. For that, for each combination of method/SNP set/GCM, we ranked the populations based on the GO predictions, i.e. populations with the lowest rank had the highest GO predictions.

### Across methods/SNP sets

We first compared the population ranks from the different combinations method/SNP set. For that, we generated one plot per GCM. We also generated a plot in which the population ranks are not based on an unique GCM but are based on the average GO predictions across all GCMs. 
In the plots below, we colored the populations that were among the three populations with the highest genomic offset in at least one combination method/SNP set. 


```{r CalcRankPop, message=FALSE}
#################################################################
# Ranking the populations for each combination method/SNP set/GCM
#################################################################

# Extracting the genomic offset predictions from LFMM, GFM and GF
# ===============================================================
sub_meth_names <- c("LFMM", "GF", "GDM")

go_df <- lapply(sub_meth_names, function(meth_name) {
  
  go_predictions <- readRDS(file=here(paste0("outputs/",meth_name,"/go_predictions.rds")))
  
  go_predictions %>% 
    lapply(function(snp_set) {

    snp_set$go %>% 
        lapply(function(gcm){
          
        tibble(pop = pop_names,
               go = gcm) %>%
            arrange(go) %>% 
            mutate(rank = rev(1:length(pop_names)))
        
      }) %>% bind_rows(.id="gcm")
        
    }) %>% bind_rows(.id="snp_set")


}) %>% 
  setNames(sub_meth_names) %>% 
  bind_rows(.id="method")


# Extracting the genomic offset predictions from the RDA and pRDA
# ===============================================================

# Which method do we use for the RDA? 
RDA_method <- "predict" # "loadings" or "predict"

# loading the genomic offset predictions from the RDA
rda_pred <- readRDS(file=here(paste0("outputs/RDA/go_predictions_",RDA_method,".rds")))

# pRDA (RDA corrected for population structure)
prda_pred <- lapply(names(snp_sets), function(set_i){
  
  rda_pred[[set_i]]$go_corrected %>% 
    
    lapply(function(gcm){ 
      
      tibble(pop = pop_names,
             go = gcm) %>%
             arrange(go) %>% 
             mutate(rank = rev(1:length(pop_names)),
                    snp_set = set_i,
                    method = "pRDA")
        
      }) %>% bind_rows(.id="gcm")
        
    }) %>% bind_rows()

# RDA (RDA not corrected for population structure)
go_df <- lapply(names(snp_sets), function(set_i){
  
  rda_pred[[set_i]]$go_uncorrected %>% 
    
    lapply(function(gcm){ 
      
      tibble(pop = pop_names,
             go = gcm) %>%
             arrange(go) %>% 
             mutate(rank = rev(1:length(pop_names)),
                    snp_set = set_i,
                    method = "RDA")
        
      }) %>% bind_rows(.id="gcm")
        
    }) %>% 
   bind_rows() %>% 
   bind_rows(prda_pred) %>% 
   bind_rows(go_df)


# ============================================================================================================
# ============================================================================================================

# For each combination of method (GF, GDM, LFMM, pRDA and RDA) and SNP set, 
# we calculate the average of the GO predictions across the five GCMs, 
# and then we rank the populations based on their GO predicted value.


# First option (that we should use I think): take the mean of the GO predictions
go_df <- go_df %>% 
  group_by(pop,method,snp_set) %>% 
  summarise(go=mean(go)) %>% 
  ungroup() %>% 
  group_split(method,snp_set) %>% 
  lapply(function(x){
    x %>% 
      arrange(go) %>% 
      mutate(rank = rev(1:length(pop_names)),
             gcm = "GCMs_average")
  }) %>% 
  bind_rows() %>% 
  bind_rows(go_df)

# Another option (the first I use, but I think we should use the option above): take the mean of the ranks
# go_df %>%
#   group_by(pop,method,snp_set) %>%
#   summarise(mean_rank=mean(rank),go=mean(go)) %>%
#   ungroup() %>%
#   group_split(method,snp_set) %>%
#   lapply(function(x){
#     x %>%
#       arrange(mean_rank) %>%
#       mutate(rank = 1:length(pop_names),
#              gcm = "GCM average") %>%
#       dplyr::select(-mean_rank)
#   }) %>%
#   bind_rows() %>%
#   bind_rows(go_df)


# =====================================
# =====================================

# FIGURE TITLE AND LABELS

# We create a vector that we will use for the x-axis labels and the plot title
go_df <- lapply(snp_sets, function(set_i){
  lapply(meth_names, function(meth_i){
    tibble(method_snpset_names = paste0(meth_i, " - ", set_i$set_name),
           meth_snpset = paste0(meth_i, "_", set_i$set_code),
           snpset_names = set_i$set_name,
           method = meth_i,
           snp_set = set_i$set_code)
    })
  }) %>% 
  bind_rows %>% 
  left_join(go_df, by = c("method", "snp_set"))

# save for the shiny app
saveRDS(go_df, here("shiny/PredictionVariability/go_df.rds"))
```

```{r}
# Which method do we show on the graph? 
selected_meth <- meth_names

# Which SNP set do we show? 
selected_sets <- names(snp_sets)[c(1,6)]

# Subseting the dataset to keep only the method and SNP sets that we want for the figures
sub_df <- go_df %>% 
  filter(snp_set %in% selected_sets,
         method %in% selected_meth)
```



```{r BumpChartsSplittingPerMethSnpSet, results='hide', message=FALSE, fig.width=12, fig.height=7}
# =================================
# Code the generate the bump charts
# =================================

# the colors I will use to color the populations with high GO
my_colors <- c(brewer.pal(n=12, "Paired"),"#FF40EE","cyan","yellow")

bump_charts <- sub_df %>% 
  #filter(! (snp_set == "cand_corrected")) %>% 
  # mutate(x_axis = method_snpset_names[paste0(method, "_", snp_set)]) %>%
  group_split(gcm) %>% 
  lapply(function(x){
    
high_go_pops <- x %>% filter(rank < 4) %>% pull(pop) %>% unique()    

my_palette <- c(my_colors[1:length(high_go_pops)],"#E8E8E8")

if(unique(x$gcm)=="GCMs_average"){
  plot_title <- "Average across the five GCMs"
} else {
  plot_title <- unique(x$gcm)
}

sub <- x %>% 
  mutate(flag = ifelse(pop %in% high_go_pops, TRUE, FALSE),
         pop_col = if_else(flag == TRUE, pop, "Others")) %>% 
  mutate(pop = factor(pop, levels=c(setdiff(unique(x$pop),high_go_pops),high_go_pops)),
         pop_col = factor(pop_col, levels = c(high_go_pops,"Others")))

p <- sub %>% ggplot(aes(x = method_snpset_names, y = rank, group = pop)) +
  geom_point(aes(color = pop_col), size = 2, alpha = 0.9) +
  geom_line(aes(color = pop_col), linewidth = 2, alpha = 0.8) +
  scale_y_reverse(breaks = 1:nrow(sub)) +
  scale_color_manual(values = my_palette) +
  geom_text(data = sub %>% filter(method_snpset_names == last(levels(factor(sub$method_snpset_names))) & 
                                    pop %in% high_go_pops),
            aes(label = pop, x = length(unique(sub$method_snpset_names)) + 0.1) , 
            hjust = 0.15, color = "gray20", size = 3) +#color = "#888888",
  theme_bw() +
  ylab("Population rank") +
  ggtitle(plot_title) +
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "none",
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=11, angle = 60, hjust = 1),
        axis.title.y = element_text(size=16),
        axis.title.x = element_blank())

ggsave(p, filename = here(paste0("figs/PredictionVariability/BumpChart_PopRank_",unique(x$gcm),".pdf")), device = "pdf", width=11, height = 7)

if(unique(x$gcm)=="GCMs_average"){
  p_manuscript <- p + ggtitle("") 
  ggsave(p_manuscript, 
         filename = here(paste0("figs/PredictionVariability/BumpChart_PopRank_",unique(x$gcm),"_notitle.pdf")), device = "pdf", width=13)} 

p

  })

pdf(here("figs/PredictionVariability/BumpCharts_PerMethSnpSet.pdf"), width=11, height = 7)
lapply(bump_charts, function(x) x)
dev.off()

bump_charts
```


## Across GCMs

We then compared the population ranks from the different GCMs (or from the average GO predictions across the five GCMs). For that, we generated one plot per combination method/SNP set. We also generated a plot in which the population ranks are not based on an unique combination method/SNP set but are based on the average GO predictions across all combinations method/SNP set.

In the plots below, we colored the populations that were among the three populations with the highest genomic offset in at least one GCM. 

```{r BumpChartsSplittingPerGCM, fig.width=10, fig.height=7}
bump_charts <- sub_df %>% 
  group_by(pop, gcm) %>% 
  summarise(go = mean(go)) %>% 
  ungroup() %>% 
  group_split(gcm) %>% 
  lapply(function(x){
    x %>% 
      arrange(go) %>% 
      mutate(rank = rev(1:length(pop_names)))
  }) %>% 
  bind_rows() %>% 
  mutate(meth_snpset = "method_snpset_average",
         method_snpset_names = "Average across methods and SNP sets") %>% 
  bind_rows(sub_df) %>% 
  group_split(meth_snpset) %>% 
  lapply(function(x){

high_go_pops <- x %>% 
  filter(rank < 4) %>% 
  pull(pop) %>% 
  unique()    

my_palette <- c(my_colors[1:length(high_go_pops)],"#E8E8E8")

df <- x %>% 
  mutate(flag = ifelse(pop %in% high_go_pops, TRUE, FALSE),
         pop_col = if_else(flag == TRUE, pop, "Others")) %>% 
  mutate(pop = factor(pop, levels=c(setdiff(unique(sub_df$pop),high_go_pops),high_go_pops)),
         pop_col = factor(pop_col, levels = c(high_go_pops,"Others")))


p <- df %>% 
  ggplot(aes(x = gcm, y = rank, group = pop)) +
  geom_point(aes(color = pop_col), size = 2, alpha = 0.9) +
  geom_line(aes(color = pop_col), linewidth = 2, alpha = 0.8) +
  scale_y_reverse(breaks = 1:nrow(sub_df)) +
  scale_color_manual(values = my_palette) +
  geom_text(data = df %>% filter(gcm == "UKESM1-0-LL" & pop %in% high_go_pops),
            aes(label = pop, x = 6.1) , hjust = 0.15, color = "gray20", size = 5) +#color = "#888888",
  theme_bw() +
  ylab("Population rank") +
  ggtitle(df$method_snpset_names) +
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.y = element_text(size=12),
        axis.text.x = element_text(size=12),
        axis.title.y = element_text(size=16),
        legend.position = "none",
        axis.title.x = element_blank())

ggsave(p, filename = here(paste0("figs/PredictionVariability/BumpChart_PopRank_",unique(x$meth_snpset),".pdf")), device = "pdf", width=10)

p
  })

pdf(here("figs/PredictionVariability/BumpCharts_PerGCM.pdf"), width=10)
lapply(bump_charts, function(x) x)
dev.off()

bump_charts
```

# Correlations

```{r CorrelationsPlots, fig.width=7, fig.height=7, results='hide'}
sub_df %>% 
  group_split(gcm) %>% 
  lapply(function(x){
    
    df_cor <- x %>% 
      pivot_wider(id_cols = pop, values_from = go, names_from = method_snpset_names)  %>% 
      dplyr::select(-pop) %>% 
      cor()
    
    
    pdf(here(paste0("figs/PredictionVariability/CorrelationPlot_",paste0(unique(x$gcm)),".pdf")))
    df_cor  %>% corrplot(method = 'number',type = 'lower', diag = FALSE,
             title="",
             mar=c(0,0,0,0),
             number.cex=1,
             tl.cex=1)
    dev.off()
      
      
    df_cor %>% corrplot(method = 'number',type = 'lower', diag = FALSE,
               title=paste0(unique(x$gcm)),mar=c(0,0,2,0),
               number.cex=1,
               tl.cex=1)
  })
```


# Maps

```{r eval=F}
list_highpops <- list(LFMM = readRDS(here("outputs/LFMM/high_go_pops.rds"))[[2]],
                      RDA = readRDS(here("outputs/RDA/high_go_pops.rds"))[[2]],
                      GDM = readRDS(here("outputs/GDM/high_go_pops.rds"))[[2]],
                      GF = readRDS(here("outputs/GF/high_go_pops.rds"))[[2]])

list_highpops <- list(LFMM = readRDS(here("outputs/LFMM/high_go_pops.rds"))[[2]],
                      RDA = readRDS(here("outputs/LFMM/high_go_pops.rds"))[[2]],
                      GDM = readRDS(here("outputs/LFMM/high_go_pops.rds"))[[2]],
                      GF = readRDS(here("outputs/LFMM/high_go_pops.rds"))[[2]])

list_highpops[[1]] <- list_highpops[[1]] + theme(legend.position = "none")
list_highpops[[2]] <- list_highpops[[2]] + theme(legend.position = "none")
list_highpops[[3]] <- list_highpops[[3]] + theme(legend.position = "none")

plot_grid(plotlist = list_highpops, nrow=2)
```



