---
title: "Validating genomic offset predictions with mortality rates in National Forest Inventory (NFI) plots"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 10px
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 5,fig.height = 4,cache=F)
options(width = 300)
library(knitr)
library(tidyverse)
library(readxl)
library(xtable)
library(reshape2)
library(kableExtra)
library(here)
library(rstan)
library(tidybayes)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(parallel)
library(geodist)
library(janitor)
library(latex2exp)
library(magrittr)
library(ggrepel)
library(paletteer)
library(cowplot)
library(ggpubr)

# Mapping
library(rnaturalearth)
world <- ne_countries(scale = "medium", returnclass = "sf")

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
source(here("scripts/functions/extract_climatedt_metadata.R")) # extracting meta data of the climatic variables in ClimateDT
```



# Introduction


**Goal:** Estimating the association between genomic offset (GO) predictions and mortality rates in natural populations of maritime pine (i.e., National Forest Inventory plots).

Mortality rates in natural populations are considered as a proxy of the absolute fitness of the populations. A positive association between GO predictions and the mortality rates would suggest that GO captures maladaptation in the face of demographic complexity (i.e., the effects of other processes on spatial variance in allele frequencies, such as expansion history and gene flow) as well as the genetic architecture of climate adaptation (polygenic trait architectures, G×E interactions, non-additive genetic variance).

We use mortality data from the French and Spanish National Forest Inventories (NFI) harmonized in @changenet2021occurrence. The French data relies on temporary plots sampled between 2005 and 2014 while the Spanish data relies on permanent plots sampled during the second (from 1986 to 1996) and third NFIs (from 1997 to 2008). A tree was recorded as dead if its death was dated at less than 5 years ago in the French NFI or if it was alive in the second inventory but dead in the third one in the Spanish NFI. 

The different census intervals among the plots are accounted for by modelling the proportion $p_{i}$ of maritime pines that died in the plot $i$ during the census interval $\Delta_i$ with the complementary log-log link and an offset on the logarithm of the census interval $\Delta_{i}$ for the plot $i$. The modeling approach was evaluated with simulations in the report `14_ValidationNFI_ModelComparison.qmd`, in which the estimation accuracy of different models was compared.

In the present report, we evaluate the relationship between mortality rates and G0 predictions with the two following models.


## Model with competition as covariate


This model corresponds to the model M3 in the report `14_ValidationNFI_ModelComparison.html`.

\begin{align*}
m_i &\sim \text{Binomial}(N_i,p_i)\\
\text{log}(-\text{log}(1-p_i)) &= \beta_{0,c} +  \beta_{C} C_i + \beta_{GO} GO_i + \text{log}(\Delta_i) \\
\end{align*}


  - $N_{i}$ is the total number of maritime pines in the plot $i$,

  - $m_{i}$ is the number of maritime pines that died during the census interval $\Delta_{i}$ in the plot $i$,
  
  - $\beta_{0,c}$ are country-specific intercepts that account for the methodological differences between the French and Spanish inventories that may bias the estimations,
  
  - $C_{i}$ is the basal area of all tree species confounded in the plot $i$ (to account for the competition among trees), with its associated coefficient $\beta_{C}$,
  
  - $X_{i}$ is the climatic distance (between the reference climate and the climate of the inventory period) or genomic offset predicted in the plot $i$, with its associated coefficient $\beta_{X}$. Note that in the scripts this variable is noted as $GO_i$, and its regression coefficient $\beta_{GO}$, because we initially did not include the climatic distances as predictors.
  
  
## Model with competition and tree age as covariates

This model corresponds to the model M6_2 in the report `14_ValidationNFI_ModelComparison.html`.


\begin{align*}
m_i &\sim \text{Binomial}(N_i,p_i)\\
\text{log}(-\text{log}(1-p_i)) &= \beta_{0,c} +  \beta_{C} C_i + \beta_{DBH} DBH_i + \beta_{C*DBH} C_i DBH_i + \beta_{X} X_i + \text{log}(\Delta_i) \\
\end{align*}


  - $DBH_{i}$ the mean diameter at breast height (DBH) of the maritime pines in the plot $i$ (including adults, dead trees and recruitment trees) and its associated coefficient $\beta_{DBH}$. This variable is expected to account for the different mean tree ages across plots that may impact mortality rates. 

We present the results from this model in the main manuscript.

## Priors

We used the following weakly informative priors:

$$\begin{bmatrix}  \beta_{0,c} \\ \beta_C \\ \beta_{X} \\ \beta_{DBH} \\ \beta_{C*DBH} \end{bmatrix} \sim \text{Normal}(0,1)$$

# NFI data


NFI data was harmonized in @changenet2021occurrence.

```{r LoadDataNFI}
nfi_data <- readRDS(here::here("data/NFIdata/NFIrawdata_Changenet2021.rds")) %>% 
  droplevels() %>% 
  as_tibble() %>% 
  dplyr::select(plotcode,
                longitude,
                latitude,
                country,
                yearsbetweensurveys, # number of years between surveys
                surveydate1, # date first survey Spanish NFI
                surveydate, # year of the second survey in the Spanish NFI and the unique survey in the French NFI
                treeNbrJ.M, # number of dead trees in the plot
                treeNbrJ.IMall, # total number of trees in the plot
                BA.ha.plot.1, # basal area of all tree species in the plot
                dbhJ.IMall.plot.mean # mean DBH of the maritime pines in the plot including dead trees but also saplings (recruitment) 
                ) %>% 
  dplyr::rename(nb_years = yearsbetweensurveys,
                nb_dead = treeNbrJ.M,
                nb_tot = treeNbrJ.IMall,
                basal_area = BA.ha.plot.1,
                mean_DBH = dbhJ.IMall.plot.mean,
                first_survey = surveydate1,
                second_survey = surveydate) %>% 
  dplyr::mutate(first_survey = case_when(!is.na(first_survey) ~ str_sub(first_survey,1,4) %>% as.numeric()), # keep only the year of the 1st survey
                prop_dead = nb_dead/nb_tot,  # proportion of dead trees
                annual_prop_dead = (nb_dead/nb_tot)/nb_years) # annual proportion of dead trees
```

Variables in the dataset

  - Plot code: `plotcode`
  - Longitude and latitude of the plot: `longitude` and `latitude`.
  - Country in which the plot is: `country` (ES = Spain, FR = France).
  - The number of years between surveys in the Spanish inventory (which is equal to 5 in the French inventory as mortality is estimated in the five years before the survey date): `nb_years`.
  - The dates of the first and second survey in the Spanish inventory: `surveydate1` and `surveydate2`.
  - The year of the second survey in the Spanish inventory and of the unique survey in the French inventory: `surveydate`.
  - The number of dead trees in the plot: `nb_dead`.
  - The total number of trees in the plot: `nb_tot`.
  - The basal area of all tree species in the plot (proxy of the competition among trees): `basal_area`.
  - The mean DBH of the maritime pines in the plot including dead trees but also saplings (recruitment): `mean_DBH`.


`r nrow(nfi_data)` plots in this dataset: `r nfi_data %>% dplyr::filter(country=='FR') %>% nrow()` from the French NFI and `r nfi_data %>% dplyr::filter(country=='ES') %>% nrow()` from the Spanish inventory.


## Filtering
 
How many NAs in each column? 

```{r NumberNANFIrawdata}
nfi_data %>% 
  summarise(across(everything(), ~ sum(is.na(.)))) %>% 
  pivot_longer(everything(), names_to = "Variable", values_to = "Number of NAs") %>% 
  kable_mydf(boldfirstcolumn = T)
```

We extract the plots with missing data for the proportion of dead trees and we look at the first eight plots.

```{r PlotsWithNaForPropDeadTrees}
df_na_prop_dead <- nfi_data %>% 
  dplyr::filter(is.na(prop_dead))
df_na_prop_dead[1:8,]  %>% kable_mydf(boldfirstcolumn = F)
```

No maritime pine was recorded in these eight plots (`nb_dead` = 0 and `nb_tot` = 0). We check that this is the case for the `r nrow(df_na_prop_dead)` plots by looking at the unique values in the columns `nb_dead` and `nb_tot` in the subset of plots with missing data for the proportion of dead trees (the `df_na_prop_dead` dataset).

```{r UniqueValuesPlotsWithNaForPropDeadTrees}
df_na_prop_dead %>% 
  summarise(across(nb_dead:nb_tot, ~ unique(.))) %>% 
  pivot_longer(everything(), names_to = "Variable", values_to = "Unique values") %>% 
  kable_mydf(boldfirstcolumn = T)
```

We remove these `r nrow(df_na_prop_dead)` plots.

```{r RemovePlotsWithNoMaritimePine}
nfi_data <- nfi_data %>% 
  dplyr::filter(!is.na(prop_dead))
```


We check that there are no more plots without maritime pine.

```{r CheckingPlotsWithoutMaritimePine}
nfi_data %>% dplyr::filter(nb_tot==0) %>% nrow()
```

Missing data for basal area:

```{r CheckingMissingDataBasalArea}
nfi_data %>% 
  summarise('Number of NAs for basal area' = sum(is.na(basal_area))) %>% 
  kable_mydf(boldfirstcolumn = F)
```

Since there are not many plots with missing basal area data, we delete them.

```{r RemovingMissingDataBasalArea}
nfi_data <- nfi_data %>% 
  dplyr::filter(!is.na(basal_area))
```

After these filtering steps, there are `r nrow(nfi_data)` plots left.


<span style="color: orange;">**DRYAD REPOSITORY:** The dataset is saved in the DRYAD repository: `NFIdata_cleaned.csv` dataset.</span>


```{r ExportNFIdataForModelComparisonAndDryadRepo}
nfi_data %>% write_csv(here("data/DryadRepo/NFIdata_cleaned.csv"))
```


## Inventory dates

```{r InventoryDates, results="hide", message=F, warning=F}
country_names <- c(`ES` = "SPAIN",`FR` = "FRANCE")

ggplot(nfi_data, aes(x=second_survey, fill=as.factor(country))) + 
  geom_histogram(alpha=0.5, position="identity") + 
  theme_bw()  +
  facet_wrap(~as.factor(country), labeller = as_labeller(country_names)) + 
  labs(y="Count of plots",x="Inventory date") +
  theme(legend.position = "none")

ggplot(nfi_data, aes(x=first_survey)) + 
  geom_histogram(alpha=0.5, position="identity") + 
  theme_bw()

nfi_data %>% 
  ggplot(aes(x=nb_years, fill=country)) + 
  geom_histogram(alpha=0.5, position="identity") + 
  labs(y="Count of plots",x="Number of years between surveys") +
  theme_bw()
```


## viz the annual prob of mortality

```{r ObservedAnnualProbMortalityMap, fig.width=6, fig.height=6}
p <- ggplot() + 
    geom_sf(data = world, fill="gray98") + 
    theme_bw() +
    scale_x_continuous(limits = c(-10, 11)) +
    scale_y_continuous(limits = c(35, 51)) + 
    geom_point(data=nfi_data, aes(x=longitude,y=latitude,color=annual_prop_dead), size=1, alpha = 0.8) + 
  xlab("") + ylab("") +
    ggtitle(NULL) +
    theme(legend.position = c(0.8,0.2),
          axis.text = element_text(size=14),
          legend.box.background = element_rect(fill = "white", linewidth=0.6, colour="gray")) +
  geom_point(data=nfi_data %>% filter(annual_prop_dead>0.07),
             aes(x=longitude,y=latitude,color=annual_prop_dead), size=1.4, alpha = 0.3) +
  geom_point(data=nfi_data %>% filter(annual_prop_dead>0.13),
             aes(x=longitude,y=latitude,color=annual_prop_dead), size=1.4, alpha = 0.3) +
  scale_color_gradientn(name = "Observed annual\nproportion of mortality", 
                        colours = c("#00FF66", "#0066FF",rep("#CCFF00",2),rep("orange",3),rep("#FF0000",5)), limits=NULL)

ggsave(plot = p, filename = here("figs/ValidationNFI/ObservedAnnualMortalityProportionMap.pdf"), height = 6, width = 6)

p
```


# Climatic data

## Generating file for ClimateDT

```{r ExportNFIplotsCoordinatesForClimateDT, eval=F}
# NFI plots coordinates
# ---------------------
nfi_climateDT <- nfi_data %>% 
  dplyr::select(plotcode, longitude, latitude) %>% 
  write_csv(here::here("data/nfi_coordinates.csv"))
```

We are going to extract the climatic data between 10 years before the first survey until the date of the second survey. For the French NFI, as there is only one survey, we will extract climatic data for the 15 years before the survey.


```{r RangeExtractionYears}
nfi_data %>%
  group_by(country) %>%
  mutate(min_year = case_when(country=="FR" ~ second_survey-15,
                              country=="ES" ~ first_survey-10)) %>% 
  summarise(min_year=min(min_year),
            max_year=max(second_survey)) %>% 
  kable_mydf()
```

So, we need annual climatic data at the location of the NFI plots for the period 1972-2014.

## Calculate climate for each plot


Once we have the annual climatic data at the location of the NFI plots, we generate two datasets:

  - one with the past climate data of the NFI plots, across the period 1901-1950.
  
  - one with the climatic data across a period specific to each plot, corresponding to:
  
    - The period from **5 years before the first inventory date** to the **second inventory date** for the Spanish NFI.
      
    - The period from **10 years before the inventory date** and the **inventory date** for the French NFI. A tree is considered dead in the French NFI if its death is estimated as less than 5 years before the inventory date. 
        
We included the **5 years preceding the estimated date of tree death**, as the climate of the previous years may have **lag effects** on tree death.


We load the annual climatic data, which corresponds to point estimate at the location of the NFI plots:

```{r LoadClimaticData}
clim <- read_csv(here("data/ClimaticData/NFIplots/ClimateDT_cmip6_GFDL-ESM4_NFIcoordinates_ADJ.csv"), show_col_types = FALSE) %>% 
  dplyr::rename(plotcode=ID,
                longitude=Longitude,
                latitude=Latitude,
                elevation=Elevation,
                year=Year)
```

We calculate the average climatic conditions across the period 1901-1950:

```{r CalcClimRef, warning=F, message=F}
# Function to calculate the average of the climatic variables at the location of the NFI plots
source(here("scripts/functions/calc_avg_clim_var.R"))

clim_ref <- clim %>% calc_avg_clim_var(ref_period = c(1901,1950), id_spatial_points = "plotcode")
```


```{r ComparingWithCoordinatesClimateDT, eval=F}
# !!Coding warning!! 
######################

# The coordinates from ClimateDT are not exactly the same as the original ones. 
# Therefore, merging the climatic data with the original data has to be done with the `plotcode` column 
# and not the latitude and longitude of the populations.

# Comparing original coordinates with the ones from ClimateDT
comp_coord <- clim_ref %>% 
  dplyr::select(plotcode,latitude,longitude) %>% 
  dplyr::rename(latitude_climDT=latitude, longitude_climDT=longitude)
comp_coord <- nfi_data %>% 
  dplyr::select(plotcode,latitude,longitude) %>% 
  inner_join(comp_coord, by="plotcode") %>% 
  mutate(diff_latitude=latitude_climDT - latitude,
         diff_longitude=longitude_climDT - longitude) %>% 
  dplyr::filter(diff_latitude != 0 | diff_longitude != 0)

# There are very very small differences
range(comp_coord$diff_latitude)
range(comp_coord$diff_longitude)
```

```{r TimePeriodsForRasterExtraction, eval=F}
# checking time periods for Maurizio
# ==================================

time_periods <- clim_survey %>% 
  dplyr::select(min_year_clim,second_survey) %>% 
  group_by(min_year_clim,second_survey) %>% 
  summarise(count=n()) 

time_periods %>% 
  print(n=nrow(.)) %>% 
  kable_mydf()

time_periods %>% filter(count <10)

time_periods %>% 
  write_csv(here("data/ClimaticData/NFIplots/NFIplots_TimePeriods.csv"))
```

We then calculate the average climatic conditions across the survey periods specific to each NFI plot:

```{r CalculateMeanClimateDuringSurveyPeriod}
# We calculate the mean climatic values between:
  # - five years before the first inventory date
  # - the second inventory date
clim_survey <- nfi_data %>% 
  mutate(min_year_clim = case_when(country=="FR" ~ second_survey-10,
                                   country=="ES" ~ first_survey-5))

# we calculate the average climatic conditions for each specific period
clim_survey <- clim_survey %>% 
  group_by(min_year_clim,second_survey) %>% 
  group_split() %>% 
  purrr::map(\(x){
  
min_year <- unique(x$min_year_clim)
max_year <- unique(x$second_survey)

clim %>% 
  right_join(x[,"plotcode"],by=c("plotcode")) %>%
  calc_avg_clim_var(ref_period = c(min_year,max_year), id_spatial_points = "plotcode")
  
}) %>% list_rbind() 


# we save in a list the average climatic conditions across the two periods (reference period and period of the NFI surveys)
list(clim_ref = clim_ref,
     clim_survey = clim_survey) %>% 
  saveRDS(here("data/ClimaticData/NFIplots/NFIclimate.rds"))
```

<span style="color: orange;">**DRYAD REPOSITORY:** We save the averaged climatic data at the locations of the NFI plots in the DRYAD repository: `ClimateDT_NFIPlots_PastClimates.csv` dataset for the averaged climates over the reference period 1901-1950 and `ClimateDT_NFIPlots_SurveyClimates.csv"` dataset for the averaged climates over the survey periods specific to each plot.</span>


```{r SaveClimateDataInDryadRepo}
clim_ref %>% write_csv(here("data/DryadRepo/ClimateDT_NFIPlots_PastClimates.csv"))
clim_survey %>% write_csv(here("data/DryadRepo/ClimateDT_NFIPlots_SurveyClimates.csv"))
```



## Climatic space covered

We look at how the 34 sampled populations cover the climatic space of the NFI plots.

```{r PlottingClimaticSpace, fig.width=8, fig.height=8}
# Selected climatic variables
clim_var <- readRDS(here("data/ClimaticData/NamesSelectedVariables.rds"))

# Reference climate of the NFI plots
clim_nfi <- clim_ref %>% 
  dplyr::select(plotcode,contains("ude"),elevation, all_of(clim_var)) %>% 
  dplyr::mutate(subdf="nfi") %>% 
  dplyr::rename(id=plotcode)

# Gene pools of the populations
gps <- readRDS(here("data/GenomicData/MainGenePoolInformation.rds"))[[1]] %>% arrange(pop)

# Reference climate of the populations
clim_pop <- readRDS(here("data/ClimaticData/MaritimePinePops/ClimatePopulationLocationPointEstimates_ReferencePeriods_noADJ.rds"))[[1]]$ref_means %>% 
  dplyr::select(pop,contains("ude"),elevation, all_of(clim_var)) %>% 
  left_join(gps,by="pop") %>% 
  dplyr::rename(id=pop,
                subdf=main_gp_pop) 

# Combining the dataset of the populations and the NFI plots
df<- bind_rows(clim_pop,clim_nfi) %>% 
  mutate(sudf=factor(subdf,levels=c("nfi",unique(clim_pop$subdf))))


# Mean annual temperature vs annual precipitation
# ===============================================
p <- df %>%
  ggplot(aes(x=bio1,y=bio12,label=id)) +
  geom_point(data=df %>% filter(subdf=="nfi"), color="gray85") +
  geom_point(data=df %>% filter(!subdf=="nfi"), aes(color=subdf),size=4) +
  scale_color_manual(values=c("darkorchid3",
                              "gold2",
                              "navyblue",
                              "turquoise2",
                              "orangered3",
                              "green3")) +
  xlab("Mean annual temperature (bio1, °C)") +
  ylab("Annual precipitation (bio12, mm)") +
  geom_text_repel(data=df %>% filter(!subdf=="nfi")) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = c(0.81,0.87),
        legend.text = element_text(size=16),
        legend.background = element_blank(),
        axis.text = element_text(size=15),
        axis.title = element_text(size=18))

p %>% ggsave(filename = here("figs/ValidationNFI/ClimaticCoverage_Bio1vsBio12.pdf"),
             device="pdf",
             width = 8,
             height=8)

p


# Precipitation seasonality vs Isothermality	
# ==========================================
p <- df %>%
  ggplot(aes(x=bio15,y=bio3,label=id)) +
  geom_point(data=df %>% filter(subdf=="nfi"), color="gray85") +
  geom_point(data=df %>% filter(!subdf=="nfi"), aes(color=subdf),size=4) +
  scale_color_manual(values=c("darkorchid3",
                              "gold2",
                              "navyblue",
                              "turquoise2",
                              "orangered3",
                              "green3")) +
  xlab("Precipitation seasonality (bio15, coefficient of variation)") +
  ylab("Isothermality (bio3, index)") +
  geom_text_repel(data=df %>% filter(!subdf=="nfi")) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = c(0.18,0.15),
        legend.text = element_text(size=16),
        legend.background = element_blank(),
        axis.text = element_text(size=15),
        axis.title = element_text(size=18))

p %>% ggsave(filename = here("figs/ValidationNFI/ClimaticCoverage_Bio3vsBio15.pdf"),
             device="pdf",
             width = 8,
             height=8)

p


# Temperature seasonality and Summer heat moisture index	
# ======================================================
p <- df %>%
  ggplot(aes(x=bio4,y=SHM,label=id)) +
  geom_point(data=df %>% filter(subdf=="nfi"), color="gray85") +
  geom_point(data=df %>% filter(!subdf=="nfi"), aes(color=subdf),size=4) +
  scale_color_manual(values=c("darkorchid3",
                              "gold2",
                              "navyblue",
                              "turquoise2",
                              "orangered3",
                              "green3")) +
  xlab("Temperature seasonality (bio4, standard deviation ×100)") +
  ylab("Summer heat moisture index (SHM, °C/mm)") +
  geom_text_repel(data=df %>% filter(!subdf=="nfi")) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = c(0.18,0.89),
        legend.text = element_text(size=16),
        legend.background = element_blank(),
        axis.text = element_text(size=15),
        axis.title = element_text(size=18))

p %>% ggsave(filename = here("figs/ValidationNFI/ClimaticCoverage_Bio4vsSHM.pdf"),
             device="pdf",
             width = 8,
             height=8)

p
```

# Climatic distances

We calculate the climatic distances between the climate during the period 1901-1950 and the climate during the inventory period. The climatic distances were calculated separately for each climatic variable and we also calculated the Euclidean climatic distances based on all selected climatic variables.

```{r CalculateClimaticDistances, eval=F}
# selected climatic variables
clim_var <- readRDS(here("data/ClimaticData/NamesSelectedVariables.rds"))

# Load the climatic data of the NFI plots.
nfi_clim <- readRDS(here("data/ClimaticData/NFIplots/NFIclimate.rds"))
  
# Keep only the climatic variables of interest and scale the climatic data
source(here("scripts/functions/generate_scaled_nfi_clim_datasets.R"))
nfi_dfs <- generate_scaled_nfi_clim_datasets(clim_var, clim_ref = nfi_clim$clim_ref, clim_pred = nfi_clim$clim_survey)

clim_dist <- lapply(clim_var, function(var){(nfi_clim[[2]][[var]] - nfi_clim[[1]][[var]]) %>% abs()}) %>% 
  setNames(clim_var) %>% 
  bind_rows() %>% 
  bind_cols(nfi_clim[[1]][,1],.)

# Calculating the Euclidean distances for all climatic variables
clim_dist$EucliDist <- map_dbl(1:nrow(nfi_dfs[[1]]),
                               ~ sqrt(sum((nfi_dfs[[1]][.x,clim_var] - nfi_dfs[[2]][.x,clim_var])^2)))

clim_dist <- clim_dist %>% 
  pivot_longer(cols=c(any_of(clim_var),"EucliDist"),names_to="snp_set_code",values_to="GO") %>% 
  mutate(method = "CD",
         method_snp_set_code = paste0("CD_",snp_set_code),
         snp_set_name = case_when(snp_set_code == "bio1" ~ "Mean annual temperature (bio1, °C)",
                                  snp_set_code == "bio12" ~ "Annual precipitation (bio12, mm)",
                                  snp_set_code == "bio15" ~ "Precipitation seasonality (bio15, index)",
                                  snp_set_code == "bio3" ~ "Isothermality (bio3, index)",
                                  snp_set_code == "bio4" ~ "Temperature seasonality (bio4, °C)",
                                  snp_set_code == "SHM" ~ "Summer heat moisture index (SHM, °C/mm)",
                                  snp_set_code == "EucliDist" ~ "All climatic variables (Euclidean distance)"),
         method_snp_set_name = paste0("CD - ", snp_set_name)) 
```

# Genomic offset predictions

We load the genomic offset predictions estimated from the different GEAs.


```{r CombineAndSaveGenomicOffsetPredictions, eval=F}
# Codes of the SNP sets
snp_sets <- readRDS(here("outputs/list_snp_sets.rds"))

# Which RDA method?
RDA_method <- "predict"

# List of methods
meth_names <- c("GDM", "GF", "LFMM", "pRDA", "RDA")

df <- lapply(snp_sets, function(x){

list_snps <- list()

list_snps$GDM <- readRDS(file=here("outputs/GDM/go_predictions.rds"))[[x$set_code]][["go_nfi"]]
list_snps$GF <- readRDS(file=here("outputs/GF/go_predictions.rds"))[[x$set_code]][["go_nfi"]]
list_snps$LFMM <- readRDS(file=here("outputs/LFMM/go_predictions.rds"))[[x$set_code]][["go_nfi"]]
list_snps$pRDA <- readRDS(file=here(paste0("outputs/RDA/go_predictions_",RDA_method,".rds")))[[x$set_code]][["go_nfi_corrected"]]
list_snps$RDA <- readRDS(file=here(paste0("outputs/RDA/go_predictions_",RDA_method,".rds")))[[x$set_code]][["go_nfi_uncorrected"]]


df <- list_snps %>% 
  bind_rows() %>% 
  mutate(snp_set_code = x$set_code,
         snp_set_name = x$set_name,
         plotcode = unique(nfi_data$plotcode))

return(df)
}) %>% 
  bind_rows() %>% 
  pivot_longer(cols= meth_names, names_to = "method", values_to ="GO") %>%
  mutate(method_snp_set_code = paste0(method, "_", snp_set_code),
         method_snp_set_name = paste0(method, " - ", snp_set_name)) %>% 
  bind_rows(clim_dist)

saveRDS(df, here("outputs/ValidationNFI/go_nfi_predictions.rds"))
```

```{r LoadGOpredictions}
df <- readRDS(here("outputs/ValidationNFI/go_nfi_predictions.rds"))
```


# Regression coefficients

We estimate the regression coefficients between genomic offset predictions and mortality rates using Bayesian hierarchical models. 

```{r SettingColorsForIntervalPlots}
# Colors for the different sets of GO method (GDM, GF, LFMM and RDA) / SNP set (control and candidate SNPs, and all SNPs for LFMM)
colors_coeff <- c("#004586FF","#FF7F0FFF","#FFD320FF","#579D1CFF","#7E0021FF","#83CAFFFF","aquamarine3","#FEB5A2FF","#F02720FF","darkviolet","deeppink","cyan2","chocolate4")
```


## Model with competition as covariate

### Stan code 

Stan code of the model (model M3 in the report `14_ValidationNFI_ModelComparison.html`):

```{r CompileMortalityModelM3, message=F,warning=F}
stancode = stan_model(here("scripts/StanModels/ValidationNFI_m3.stan"))
print(stancode)
```

### Running the model

Running the model for each method (GDM, LFMM, GF and RDA) and each SNP set (all SNPs - only for LFMM-, candidate SNPs and control SNPs):

```{r RunMortalityModelsM3, eval=F}
# Parameters to estimate
params_to_estimate <- c("alpha_country","beta_GO","beta_C")

# Running one model for each of the nine sets of method / SNP set
coefftab <- lapply(unique(df$method), function(method_i){
  
# Subset the data - keeping only one set of method / SNP set
    sub_data <- df %>% 
      filter(method == method_i) %>% 
      inner_join(nfi_data, by="plotcode")
      
# Data in a list for Stan 
    stanlist <- list(N = nrow(sub_data),
                     nb_dead = sub_data$nb_dead,
                     nb_tot=sub_data$nb_tot,
                     GO = (sub_data$GO - mean(sub_data$GO))/sd(sub_data$GO),
                     C = (sub_data$basal_area - mean(sub_data$basal_area))/sd(sub_data$basal_area),
                     nb_country=length(unique(sub_data$country)),
                     country=as.numeric(as.factor(sub_data$country)),
                     log_nb_years=log(sub_data$nb_years))

       

# Running the Stan model
mod <- sampling(stancode, 
                data = stanlist, 
                iter = 2000, 
                hains = 4, 
                cores = 4,
                init=0, 
                save_warmup = FALSE,
                warmup = 1000)

# Model coefficients
conf95 <- broom.mixed::tidyMCMC(mod,
                                pars=params_to_estimate,
                                droppars = NULL, 
                                robust = FALSE, # give mean and standard deviation
                                ess = T, # effective sample size estimates
                                rhat = T, # Rhat estimates
                                conf.int = T, conf.level = 0.95 # include 95% credible intervals
                                ) %>% 
    dplyr::rename(conf95_low=conf.low,
                  conf95_high=conf.high,
                  mean=estimate,
                  std_deviation=std.error)
  
  
  broom.mixed::tidyMCMC(mod,
                        pars=params_to_estimate,
                        droppars = NULL, 
                        robust = TRUE, # give median and mean absolute deviation (= avg distance btw each data point and the mean)
                        ess = F, rhat = F, 
                        conf.int = T, conf.level = 0.80 # include 80% credible intervals
                        ) %>% 
    dplyr::rename(conf80_low=conf.low,
                  conf80_high=conf.high,
                  median=estimate,
                  mean_absolute_deviation=std.error) %>%
    inner_join(conf95,by=c("term")) %>% 
    mutate(method=method_i,
           method_type=unique(sub_data$method_type),
           method_input=unique(sub_data$method_input),
           .before=1)
    
  }) %>% bind_rows()

coefftab %>% saveRDS(file=here("outputs/ValidationNFI/ModelM3_coefftab.rds"))
```

### Model coefficients

Below, we plot the 95% credible intervals of:

  - the **$\beta_{X}$ coefficients**, which stand for the association between mortality rates in NFI plots and the genomic offset predictions or the climatic distances.
  
  - the **$\beta_C$ coefficients**, which stands for the association between the tree basal area (all species confounded) in the NFI plots and the mortality rates.


```{r IntervalPlotsM3, fig.height=5, fig.width=5, eval=T, results="hide"}
coefftab <- readRDS(file=here("outputs/ValidationNFI/ModelM3_coefftab.rds"))
  
coeff_match <- list(beta_GO=list("Regression coefficients $\\beta_{GO}$",c(0.77,0.85)),#"$\\beta_{GO}$ estimates",
                    beta_C=list("Regression coefficients $\\beta_C$",c(0.77,0.16)))

p <- lapply(c("beta_GO","beta_C"), function(coeff){
  
p <- coefftab %>% 
  filter(term==coeff) %>% 
  mutate(method_input=case_when(method_input=="all" ~ "All SNPs",
                                method_input=="cand" ~ "Candidate SNPs",
                                method_input=="control" ~ "Control SNPs")) %>% 
  mutate(method_input=factor(method_input, levels=c("All SNPs",
                                                    "Candidate SNPs",
                                                    "Control SNPs"))) %>% 
    ggplot(aes(x = method_type,
               y = median,
               ymin = conf95_low, 
               ymax = conf95_high,
               color=method_input,
               shape=method_input)) +
  
  {if(coeff=="beta_GO")  geom_rect(inherit.aes=FALSE, aes(xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf), color="transparent", fill="green", alpha=0.01)} + 
  geom_hline(yintercept = 0,color="gray") +
  geom_pointinterval(position = position_dodge(width = .4),point_size=3.5,size=3) +
  ylab(TeX(coeff_match[[coeff]][[1]])) + xlab("") +
  ylim(c(-0.15,0.23)) +
  scale_color_manual(values=colors_coeff) +
  scale_shape_manual(values = c(rep(16,6),rep(17,3))) +
  theme_bw() +
  labs(color="SNP set",shape="SNP set") +
  theme(axis.text.x =  element_text(size=13),
        axis.text.y = element_text(size=13),
        axis.title.y = element_text(size=16),
        axis.title.x = element_text(size=1),
        legend.title=element_text(size=13), 
        legend.box.background = element_rect(colour = "gray", linewidth=0.6),
        strip.text.x = element_text(size = 16),
        strip.background = element_blank(),
        legend.position = coeff_match[[coeff]][[2]],
        legend.text=element_text(size=12),
        panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank()) +
  guides(color=guide_legend(ncol=1),
         shape = guide_legend(override.aes = list(size =2 )))

# save in pdf and png
ggsave(p,file=here(paste0("figs/ValidationNFI/ModelM3_",coeff,"_IntervalPlots.pdf")),
       device="pdf",
       height=5,
       width=5)

return(p)

})

p
```

<span style="color: orange;">Interpretation</span>

As expected, the basal area is positively associated with mortality rates.

**For the four methods (GDM, LFMM, GF and RDA), populations with higher genomic offset based on the candidate SNPs (and all SNPs for LFMM) had higher mortality rates in the NFI plots**. Interestingly, this was not the case with genomic offset predictions based on the control SNPs: for GF and LFMM, populations with higher genomic offset based on the control SNPs also showed higher mortality rates but for GDM and RDA, populations with higher genomic offset showed lower mortality rates in the NFI plots. This confirms the importance of selecting a set of alleles potentially under selection before using the genomic offset predictions to predict maladaptation in natural populations (or using all SNPs with the LFMM approach). 

We can also note the very high association between mortality rates and the genomic offset predictions based on candidate SNPs and from the GF and GDM methods.

### $\beta_{GO}$ estimates against a null distribution

We evaluated the probability that the estimated relationship between the mortality rates and GO predictions could be obtained by chance. For that, we ran again the model on the NFI data but we replaced GO predictions by a randomly generated variable $X$ such as $X \sim \mathcal{N}(0,1)$. We repeated this step 100 times and obtained a distribution of $\beta_X$ coefficients capturing the relationship between mortality rates and a randomly generated variable. We then compared this distribution of $\beta_X$ coefficients to the estimated $\beta_{GO}$ coefficients capturing the relationship between the mortality rates and the GO predictions.

```{r IntervalPlotsComparingRandomVsPredictedGOM3, fig.height= 10, fig.width=8, eval=T}
# Load the coefficient values of models with predicted GO
coefftab <- readRDS(file=here("outputs/ValidationNFI/ModelM3_coefftab.rds")) %>% 
  mutate(run_ID=method,
         GO="predicted") %>% 
  dplyr::filter(term == "beta_GO") %>%
  dplyr::select(run_ID, GO, mean, contains("conf"))

# Load the coefficient values of models with random GO and merge with the predicted GO coefficient table
coefftab <- readRDS(file=here(paste0("outputs/ValidationNFI/Simulations/100simulations_m3_truedata_randomGO.rds"))) %>% 
  bind_rows(.id="sim_ID") %>% 
  dplyr::filter(term == "beta_GO") %>%
  mutate(run_ID = paste0("sim ",sim_ID),#run_ID = paste0("Simulation ",sim_ID),
         GO="random") %>% 
  dplyr::select(run_ID, GO, mean, contains("conf")) %>% 
  bind_rows(coefftab)
  
# Plot the mean and 95% credible intervals with interval plots
p <- coefftab %>% 
  ggplot(aes(mean, reorder(run_ID, mean))) +
  geom_vline(xintercept = 0, color="gray30") +
  geom_errorbar(aes(xmin = conf95_low, xmax = conf95_high, color=GO)) +
  scale_color_manual(values=c("magenta","gray"),
                     name = "Genomic offset") +
  geom_point() +
  theme_bw() +
  labs(x="GO effect size", y="") +
  theme(axis.text.y = element_text(size=6))

# save the plot in pdf
ggsave(p,
       file=here(paste0("figs/ValidationNFI/ModelM3_IntervalPlots_RandomGO_vs_PredictedGO.pdf")),
       device="pdf",
       height=10,
       width=8)

p
```

This graph shows that it is very unlikely that the relationship between the mortality rates and GO predictions of GDM + candidate SNPs, GF + candidate SNPs, RDA + candidate SNPs and GDM + control SNPs are obtained by chance.

## Model with competition and tree age as covariates

### Stan code 

Stan code of the model (model M6_2 in the report `14_ValidationNFI_ModelComparison.html`):

```{r CompileMortalityModelM62, message=F,warning=F}
stancode = stan_model(here("scripts/StanModels/ValidationNFI_m6_2.stan"))
print(stancode)

# Parameters to estimate
params_to_estimate <- c("alpha_country","beta_GO","beta_C", "beta_DBH","beta_C_DBH")
```

### Running the model

```{r RunMortalityModelsM62, eval=F}
# Running one model for each of the nine sets of method / SNP set
lapply(unique(df$method_snp_set_code), function(method_snp_set_i){

  # Subset the data - keeping only one set of method / SNP set
  sub_data <- df %>% 
    filter(method_snp_set_code == method_snp_set_i) %>% 
    inner_join(nfi_data, by="plotcode")
      
  # Data in a list for Stan 
  stanlist <- list(N = nrow(sub_data),
                   nb_dead = sub_data$nb_dead,
                   nb_tot=sub_data$nb_tot,
                   GO = (sub_data$GO - mean(sub_data$GO))/sd(sub_data$GO),
                   C = (sub_data$basal_area - mean(sub_data$basal_area))/sd(sub_data$basal_area),
                   DBH = (sub_data$mean_DBH - mean(sub_data$mean_DBH))/sd(sub_data$mean_DBH),
                   nb_country=length(unique(sub_data$country)),
                   country=as.numeric(as.factor(sub_data$country)),
                   log_nb_years=log(sub_data$nb_years))

       

  # Running the Stan model
  mod <- sampling(stancode, 
                  data = stanlist, 
                  iter = 2000, 
                  chains = 4, 
                  cores = 4,
                  init=0, 
                  save_warmup = FALSE,
                  warmup = 1000, 
                  pars = params_to_estimate,
                  verbose =T)

  # Save the model and the stanlist
  list(mod = mod, stanlist = stanlist) %>% 
    saveRDS(file=here(paste0("outputs/ValidationNFI/StanModels/M62_",method_snp_set_i,".rds")))

  })
```


```{r CoeffTablesMortalityModelsM62, eval=F}
coefftab <- lapply(unique(df$method_snp_set_code), function(method_snp_set_i){

  # Subset the data - keeping only one set of method / SNP set
  sub_data <- df %>% 
    filter(method_snp_set_code == method_snp_set_i)
  
 list_mod <- readRDS(file=here(paste0("outputs/ValidationNFI/StanModels/M62_",method_snp_set_i,".rds")))
 
 # Model coefficients
 conf95 <- broom.mixed::tidyMCMC(list_mod$mod,
                                  pars=params_to_estimate,
                                  droppars = NULL, 
                                  robust = FALSE, # give mean and standard deviation
                                  ess = T, # effective sample size estimates
                                  rhat = T, # Rhat estimates
                                  conf.int = T, conf.level = 0.95 # include 95% credible intervals
                                  ) %>% 
    dplyr::rename(conf95_low=conf.low,
                  conf95_high=conf.high,
                  mean=estimate,
                  std_deviation=std.error)
  
  broom.mixed::tidyMCMC(list_mod$mod,
                        pars=params_to_estimate,
                        droppars = NULL, 
                        robust = TRUE, # give median and mean absolute deviation (= avg distance btw each data point and the mean)
                        ess = F, 
                        rhat = F, 
                        conf.int = T, 
                        conf.level = 0.80 # include 80% credible intervals
                        ) %>% 
    dplyr::rename(conf80_low=conf.low,
                  conf80_high=conf.high,
                  median=estimate,
                  mean_absolute_deviation=std.error) %>%
    inner_join(conf95,by=c("term")) %>% 
    mutate(method_snp_set_code = method_snp_set_i,
           method_snp_set_name = unique(sub_data$method_snp_set_name),
           snp_set_name = unique(sub_data$snp_set_name),
           snp_set_code = unique(sub_data$snp_set_code),
           method = unique(sub_data$method),
           .before=1) %>% 
    saveRDS(file=here(paste0("outputs/ValidationNFI/StanModels/ModelM62_",method_snp_set_i,"_coefftab.rds")))
  
})
```


```{r MergeCoeffTablesMortalityModelsM62, eval=F}
coefftab <- lapply(unique(df$method_snp_set_code), function(method_snp_set_i){
  readRDS(file=here(paste0("outputs/ValidationNFI/StanModels/ModelM62_",method_snp_set_i,"_coefftab.rds")))
  }) %>% 
  bind_rows() 
  
coefftab %>% mutate(snp_set_name = factor(snp_set_name, levels = c(unique(coefftab$snp_set_name)[[6]],
                                                        unique(coefftab$snp_set_name)[[4]],
                                                        unique(coefftab$snp_set_name)[[5]],
                                                        unique(coefftab$snp_set_name)[[3]],
                                                        unique(coefftab$snp_set_name)[[1]],
                                                        unique(coefftab$snp_set_name)[[2]],
                                                        unique(coefftab$snp_set_name)[[7]],
                                                        unique(coefftab$snp_set_name)[[8]],
                                                        unique(coefftab$snp_set_name)[[9]],
                                                        unique(coefftab$snp_set_name)[[10]],
                                                        unique(coefftab$snp_set_name)[[11]],
                                                        unique(coefftab$snp_set_name)[[12]],
                                                        unique(coefftab$snp_set_name)[[13]])),
          method = factor(method, levels = c(unique(coefftab$method)[[1]],
                                             unique(coefftab$method)[[2]],
                                             unique(coefftab$method)[[3]],
                                             unique(coefftab$method)[[4]],
                                             unique(coefftab$method)[[5]],
                                             unique(coefftab$method)[[6]]))) %>% 
  saveRDS(file=here("outputs/ValidationNFI/ModelM62_coefftab.rds"))
```

### Predictions of tree mortality probability

#### Plots

Let's visualize the uncertainty around the estimation of the probability of tree mortality $p$.


```{r MortalityPredictions, message=F,warning=F, fig.height=12, fig.width=12, eval=F}
############################################################################################
# Visualizing the relationships between GO predictions and mortality probability predictions
############################################################################################
selected_method_snp_set <- unique(df$method_snp_set_code)[!grepl("CD", unique(df$method_snp_set_code))]

myplots <- lapply(selected_method_snp_set, function(method_snp_set_code_i){

# We load the stanlist
stanlist <- readRDS(file = here(paste0("outputs/ValidationNFI/StanModels/M62_",method_snp_set_code_i,".rds")))$stanlist

# We load the model
mod <- readRDS(file = here(paste0("outputs/ValidationNFI/StanModels/M62_",method_snp_set_code_i,".rds")))$mod

# We extract the posterior distributions of all parameters
post <- as.data.frame(mod)

# Vector of predictor values (based on the min and max of the GO predictions)
x_vec <- seq(min(stanlist$GO),max(stanlist$GO),0.05)

# Function to predict the mortality probability p with the initial tree height fixed to the mean
p_pred <- lapply(c("alpha_country[1]","alpha_country[2]"), function(country_i){
  
  f_p <- function(x) 1 -(exp(-exp(post[,country_i] + 
                                    post$`beta_DBH` * mean(stanlist$DBH) + 
                                    post$`beta_C` * mean(stanlist$C) + 
                                    post$`beta_C_DBH` * mean(stanlist$C) * mean(stanlist$DBH) + 
                                    post$`beta_GO` * x +
                                    log(5))))
  
sapply(x_vec, f_p) %>% 
    tibble::as_tibble() %>%
    rename_all(function(x) x_vec) %>%
    mutate(Iter = row_number()) %>%
    gather(x, p, -Iter) %>%
    group_by(x) %>%
    mutate(hpdi_l = HDInterval::hdi(p, credMass = 0.90)[1],
           hpdi_h = HDInterval::hdi(p, credMass = 0.90)[2],
           p_mean = mean(p)) %>%
    ungroup() %>%
    mutate(x = as.numeric(x),
           Country=ifelse(country_i=="alpha_country[1]","Spain","France"))
  
}) %>% bind_rows()


# Keep mean and 90% HDPI of p (one value for each iteration)
p_mean_df <- p_pred %>%
  dplyr::select(x,hpdi_l,hpdi_h,p_mean,Country) %>% 
  dplyr::distinct()


# Plots
#=======

# Plot options 
y_limits <- c(0,0.2)
x_axis <- "Mean-centered GO predictions"
y_axis <- "Probability of tree mortality"

p <- ggplot() +
  ylim(y_limits) +
  labs(y=y_axis, x=x_axis) +
  theme_bw() +
  theme(legend.position=c(0.2,0.85),
        legend.text = element_text(size=13),
        legend.title = element_text(size=13),
        legend.background = element_blank())

# First 100 posterior draws of the mortality probability p
p1 <- p +
  geom_point(data = p_pred %>% filter(Iter < 101),
             aes(x, p, color=Country), alpha = .2) 
  
# Mean (line) and 90% HPDI (shaded region) of the mortality probability p
p2 <- p +
  geom_ribbon(data = p_mean_df,
              aes(x = x, ymin = hpdi_l, ymax = hpdi_h, fill=Country),
              alpha = .2) +
  geom_line(data = p_mean_df,
              aes(x=x, y=p_mean,col=Country))

p1_p2 <- ggarrange(p1, p2 + labs(y=""), nrow = 1)

# Title
text_title <- df %>% filter(method_snp_set_code == method_snp_set_code_i) %>% .[["method_snp_set_name"]] %>% unique()

title <- ggdraw() + 
  draw_label(text_title,
             fontface = 'bold',
             x = 0, hjust = 0) +   
  theme(plot.margin = margin(0, 0, 0, 7))

# merge title and plots
p1_p2 <- plot_grid(title, p1_p2, ncol = 1, rel_heights = c(0.1, 1))




ggsave(p1_p2,
       file=here(paste0("figs/ValidationNFI/MortalityProbabilityPredictions_UncertaintyIntervals/M62_",method_snp_set_code_i,".pdf")),
       device="pdf",
       height=6,
       width=11)

# Run these rows if you want to build a figure with p2 plots for different method/SNP sets.
# p2 <- p2 + labs(subtitle = text_title)
# return(p2)

})

# p <- plot_grid(myplots[[1]] + labs(x="") + theme(legend.position = "none"),
#                myplots[[2]] + labs(y="",x="") + theme(legend.position = "none"),
#                myplots[[3]] + labs(y="",x="") + theme(legend.position = c(0.8,0.85)),
#                myplots[[4]] + labs(x="") + theme(legend.position = "none"),
#                myplots[[5]] + labs(y="",x="") + theme(legend.position = "none"),
#                myplots[[6]] + labs(y="",x="") + theme(legend.position = "none"),
#                myplots[[7]] + theme(legend.position = "none"),
#                myplots[[8]] + labs(y="") + theme(legend.position = "none"),
#                myplots[[9]] + labs(y="") + theme(legend.position = "none"),
#                ncol=3)
# 
# ggsave(p,
#        file=here(paste0("figs/ValidationNFI/MortalityProbabilityPredictions_M62.pdf")),
#        device="pdf",
#        height=12,
#        width=12)
# 
# p
```



#### Table

In the table below:

  - mean and 90% credible intervals (highest posterior density intervals) of $p$ for $x=\{-1, 0, 1\}$.
  
  - percent change in $p$ associated with a one-unit increase in $x$ (which corresponds to a one-standard deviation increase).

```{r MortalityPredictionsTable, message=F,warning=F, fig.height=6, fig.width=6, eval=F}
p_pred <- lapply(unique(df$method_snp_set_code), function(method_snp_set_code_i){

# We load the stanlist
stanlist <- readRDS(file = here(paste0("outputs/ValidationNFI/StanModels/M62_",method_snp_set_code_i,".rds")))$stanlist

# We load the model
mod <- readRDS(file = here(paste0("outputs/ValidationNFI/StanModels/M62_",method_snp_set_code_i,".rds")))$mod

# We extract the posterior distributions of all parameters
post <- as.data.frame(mod)

# Vector of predictor values (based on the min and max of the GO predictions)
x_vec <- c(-1,0,1)

# Function to predict the mortality probability p with the initial tree height fixed to the mean
p_pred <- lapply(c("alpha_country[1]","alpha_country[2]"), function(country_i){
  
  f_p <- function(x) 1 -(exp(-exp(post[,country_i] + 
                                    post$`beta_DBH` * mean(stanlist$DBH) + 
                                    post$`beta_C` * mean(stanlist$C) + 
                                    post$`beta_C_DBH` * mean(stanlist$C) * mean(stanlist$DBH) + 
                                    post$`beta_GO` * x +
                                    log(5))))
  
sapply(x_vec, f_p) %>% 
    tibble::as_tibble() %>%
    rename_all(function(x) x_vec) %>%
    mutate(Iter = row_number()) %>%
    gather(x, p, -Iter) %>%
    group_by(x) %>%
    mutate(hpdi_l = HDInterval::hdi(p, credMass = 0.90)[1],
           hpdi_h = HDInterval::hdi(p, credMass = 0.90)[2],
           p_mean = mean(p)) %>%
    ungroup() %>%
    mutate(x = as.numeric(x),
           Country=ifelse(country_i=="alpha_country[1]","Spain","France"))
  
}) %>%  
  bind_rows() %>% 
  dplyr::select(x,p_mean,hpdi_l,hpdi_h,Country) %>% 
  distinct() %>% 
  mutate_if(is.numeric, round,  digits=3)

lapply(unique(p_pred$Country), function(country_i){
  
p_pred_sub <- p_pred %>% dplyr::filter(Country==country_i)

delta_p_pos <- (p_pred_sub$p_mean[p_pred_sub$x==1] * 100 / p_pred_sub$p_mean[p_pred_sub$x==0])-100
delta_p_neg <- (p_pred_sub$p_mean[p_pred_sub$x==-1] * 100 / p_pred_sub$p_mean[p_pred_sub$x==0])-100

tibble("Method" = df %>% filter(method_snp_set_code == method_snp_set_code_i) %>% .[["method_snp_set_name"]] %>% unique(),
       "Country"=country_i,
         "p(X=-1)"=paste0(p_pred_sub$p_mean[p_pred_sub$x==-1], " [",
                       p_pred_sub$hpdi_l[p_pred_sub$x==-1],";",
                       p_pred_sub$hpdi_h[p_pred_sub$x==-1],"]"),
       "p(X=0)"=paste0(p_pred_sub$p_mean[p_pred_sub$x==0], " [",
                       p_pred_sub$hpdi_l[p_pred_sub$x==0],";",
                       p_pred_sub$hpdi_h[p_pred_sub$x==0],"]"),
       "p(X=1)"=paste0(p_pred_sub$p_mean[p_pred_sub$x==1], " [",
                       p_pred_sub$hpdi_l[p_pred_sub$x==1],";",
                       p_pred_sub$hpdi_h[p_pred_sub$x==1],"]"),
       "+Δp"=paste0(round(delta_p_pos,3),"%"),
       "-Δp"=paste0(round(delta_p_neg,3),"%"))
  
}) %>% bind_rows()

})  %>% bind_rows()

p_pred %>% saveRDS(file = here("tables/MortalityPredNFI.rds"))
```

```{r ShowMortalityPredictionsTable}
readRDS(file = here("tables/MortalityPredNFI.rds"))  %>% kable_mydf()
```


### Model coefficients

Below, we plot the 95% credible intervals of:

  - the **$\beta_{X}$ coefficients**, which stand for the association between mortality rates in NFI plots and the genomic offset predictions (i.e. capturing the potential maladaptation of the populations at the location of the NFI plots).
  
  - the **$\beta_C$ coefficients**, which stands for the association between the tree basal area (all species confounded) in the NFI plots and the mortality rates.
  
  - the **$\beta_{DBH}$ coefficients**, which stands for the association between the mean DBH (diameter at breast height) of all maritime pines in the NFI plots (including adults, dead trees and seedlings/saplings) and the mortality rates.
    
  - the **$\beta_{C*DBH}$ coefficients**, which stands for the interaction between the mean DBH and the tree basal area.


```{r IntervalPlotsM62, fig.height=8, fig.width=10, eval=T, results="hide"}
coefftab <- readRDS(file=here("outputs/ValidationNFI/ModelM62_coefftab.rds"))
  
coeff_match <- list(beta_GO=list("Regression coefficients $\\beta_{X}$ in mortality models",
                                 c(0.5,0.85),
                                 c(-0.2,0.4)), 
                    beta_C=list("Regression coefficients $\\beta_C$ in mortality models",
                                c(0.5,0.85),
                                c(-0.01,0.2)),
                    beta_DBH=list("Regression coefficients $\\beta_{DBH}$ in mortality models",
                                  c(0.5,0.85),
                                  c(-0.05,0.45)),
                    beta_C_DBH=list("Regression coefficients $\\beta_{C*DBH}$ in mortality models",
                                    c(0.5,0.85),
                                    c(-0.2,0.1)))

p <- lapply(c("beta_GO","beta_C","beta_DBH","beta_C_DBH"), function(coeff){
  
p <- coefftab %>% 
  filter(term==coeff) %>% 
  ggplot(aes(x = method,
               y = median,
               ymin = conf95_low, 
               ymax = conf95_high,
               color = snp_set_name,
               shape = snp_set_name)) +
  {if(coeff=="beta_GO")  geom_rect(inherit.aes=FALSE, aes(xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf), color="transparent", fill="green", alpha=0.005)} + 
  geom_hline(yintercept = 0,color="gray") +
  geom_pointinterval(position = position_dodge(width = .4),point_size=3.5,size=3) +
  ylab(TeX(coeff_match[[coeff]][[1]])) + xlab("") +
  ylim(coeff_match[[coeff]][[3]]) +
  scale_color_manual(values=colors_coeff) +
  scale_shape_manual(values = c(rep(16,6),rep(17,7))) +
  theme_bw() +
  labs(color="",shape="") +
  theme(axis.text.x =  element_text(size=17),
        axis.text.y = element_text(size=17),
        axis.title.y = element_text(size=19),
        legend.title=element_blank(), 
        legend.box.background = element_rect(colour = "gray", linewidth=0.6),
        strip.text.x = element_text(size = 16),
        strip.background = element_blank(),
        legend.position = coeff_match[[coeff]][[2]],
        legend.text=element_text(size=11),
        panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank()) +
  guides(color=guide_legend(ncol=2),
         shape = guide_legend(override.aes = list(size =2 )))

# save in pdf and png
ggsave(p,file=here(paste0("figs/ValidationNFI/ModelM62_",coeff,"_IntervalPlots.pdf")),
       device="pdf",
       height=8.8,
       width=9)

return(p)

})

p
```


### $\beta_{GO}$ estimates against a null distribution

We evaluated the probability that the estimated relationship between the mortality rates and GO predictions could be obtained by chance. For that, we ran again the model on the NFI data but we replaced GO predictions by a randomly generated variable $X$ such as $X \sim \mathcal{N}(0,1)$. We repeated this step 100 times and obtained a distribution of $\beta_X$ coefficients capturing the relationship between mortality rates and a randomly generated variable. We then compared this distribution of $\beta_X$ coefficients to the estimated $\beta_{GO}$ coefficients capturing the relationship between the mortality rates and the GO predictions.

```{r IntervalPlotsComparingRandomVsPredictedGOM62, fig.height=12,fig.width=8, eval=T}
# Load the coefficient values of models with predicted GO
coefftab <- readRDS(file=here("outputs/ValidationNFI/ModelM62_coefftab.rds")) %>% 
  mutate(run_ID = method_snp_set_name,
         GO = "predicted") %>% 
  dplyr::filter(term == "beta_GO") %>%
  dplyr::select(run_ID, GO, mean, contains("conf"))

# Load the coefficient values of models with random GO and merge with the predicted GO coefficient table
coefftab <- readRDS(file=here(paste0("outputs/ValidationNFI/Simulations/100simulations_m6_2_truedata_randomGO.rds"))) %>% 
  bind_rows(.id = "sim_ID") %>% 
  dplyr::filter(term == "beta_GO") %>%
  mutate(run_ID = paste0("sim ",sim_ID),#run_ID = paste0("Simulation ",sim_ID),
         GO = "random") %>% 
  dplyr::select(run_ID, GO, mean, contains("conf")) %>% 
  bind_rows(coefftab)
  
# Plot the mean and 95% credible intervals with interval plots
p <- coefftab %>% 
  ggplot(aes(mean, reorder(run_ID, mean))) +
  geom_vline(xintercept = 0, color="gray30") +
  geom_errorbar(aes(xmin = conf95_low, xmax = conf95_high, color=GO)) +
  scale_color_manual(values=c("magenta","gray"),
                     name = "Predictor") +
  geom_point() +
  theme_bw() +
  labs(x="Predictor effect size", y="") +
  theme(axis.text.y = element_text(size=6),
        legend.box.background = element_rect(colour = "gray", linewidth=0.6),
        legend.position = c(0.9,0.1),
        legend.text=element_text(size=10))

# save the plot in pdf
ggsave(p,
       file=here(paste0("figs/ValidationNFI/ModelM62_IntervalPlots_RandomGO_vs_PredictedGO.pdf")),
       device="pdf",
       height=12,
       width=10)

p
```

### Mapping model predictions

We add in the Stan code of the model the estimation of:

  - $p$ the probability of mortality during the survey period in each plot (`p` in the Stan code).
  
  - $p_{annual}$ the annual probability of mortality in each plot (`p_annual` in the Stan code).
  
  - $p_{annual,go}$ the annual probability of mortality in each plot with $DBH$ and $C$ fixed to the mean and country fixed for Spain (`p_annual_go` in the Stan code).
  
  - $m_{pred}$ the predicted number of dead trees in each plot during the survey period (`nb_dead_pred` in the Stan code).
  
  - $r_{pred}$ the residuals (in each plot during the survey period): the observed number of dead trees $m_i$ minus the predicted number of dead trees $m_{pred}$ (`nb_dead_res` in the Stan code).


```{r CompileMortalityModelM62WithPredictions, message=F,warning=F}
# Stan code
stancode = stan_model(here("scripts/StanModels/ValidationNFI_m6_2_with_predictions.stan"))
print(stancode)
```

We estimate these parameters for genomic offset predictions with GF and the sets with all candidate SNPs.

```{r ChosingMethodAndSNPsetForMortalityModelM62PredictionMapping}
method_snp_set_code_i <- "GF_all_cand"
```

```{r RunMortalityModelM62WithPredictions, eval=F}
params_to_estimate <- c("p", "p_annual", "p_annual_go", "nb_dead_pred", "nb_dead_res")

# Subset the data - keeping only one set of method / SNP set
sub_data <- df %>% 
  filter(method_snp_set_code == method_snp_set_code_i) %>% 
  inner_join(nfi_data, by="plotcode")
      
saveRDS(sub_data, file = here(paste0("outputs/ValidationNFI/StanModels/M62_with_predictions/subdata_M62_with_predictions_",method_snp_set_code_i,".rds")))

# Data in a list for Stan 
stanlist <- list(N = nrow(sub_data),
                     nb_dead = sub_data$nb_dead,
                     nb_tot=sub_data$nb_tot,
                     GO = (sub_data$GO - mean(sub_data$GO))/sd(sub_data$GO),
                     C = (sub_data$basal_area - mean(sub_data$basal_area))/sd(sub_data$basal_area),
                     DBH = (sub_data$mean_DBH - mean(sub_data$mean_DBH))/sd(sub_data$mean_DBH),
                     nb_country=length(unique(sub_data$country)),
                     country=as.numeric(as.factor(sub_data$country)),
                     log_nb_years=log(sub_data$nb_years))

saveRDS(stanlist, file = here(paste0("outputs/ValidationNFI/StanModels/M62_with_predictions/stanlist_M62_with_predictions_",method_snp_set_code_i,".rds")))

# Running the Stan model
sampling(stancode, data = stanlist, iter = 1500, chains = 4, cores = 4,init=0, save_warmup = FALSE, include = TRUE, pars=params_to_estimate) %>%
  broom.mixed::tidyMCMC(pars=params_to_estimate,
                             droppars = NULL, 
                             robust = TRUE,  # give median and mean absolute deviation (= avg distance btw each data point and the mean)
                             ess = T, # effective sample size estimates
                             rhat = T, # Rhat estimates
                             conf.int = T, 
                             conf.level = 0.95 # include 95% credible intervals
                        ) %>% 
  saveRDS(file=here(paste0("outputs/ValidationNFI/StanModels/M62_with_predictions/coefftab_",method_snp_set_code_i,".rds")))
```

We map the predictions.

```{r MappingM62Predictions}
# Package for mapping
library(rnaturalearth)
world <- ne_countries(scale = "medium", returnclass = "sf")

# Function for mapping
map_pred_nfi <- function(x, # parameter to plot 
                         ggtitle = NULL, 
                         legendtitle = "Annual probability\nof tree mortality",
                         plots_above, # Which plots should be overlaid on top to be sure of seeing them?
                         method_snp_set_code,
                         unbalanced_rainbow){

coeffs <- readRDS(file=here(paste0("outputs/ValidationNFI/StanModels/M62_with_predictions/coefftab_",method_snp_set_code,".rds"))) %>% 
  filter(str_detect(term, paste0("^", x, "\\[")))
  
data <- readRDS(file=here(paste0("outputs/ValidationNFI/StanModels/M62_with_predictions/subdata_M62_with_predictions_",method_snp_set_code,".rds"))) %>% 
  dplyr::select(plotcode, longitude, latitude) %>% 
  bind_cols(coeffs) 

p <- ggplot() + 
    geom_sf(data = world, fill="gray98") + 
    theme_bw() +
    scale_x_continuous(limits = c(-10, 11)) +
    scale_y_continuous(limits = c(35, 51)) + 
    geom_point(data=data, aes(x=longitude,y=latitude,color=estimate), size=1, alpha = 0.8) + 
  xlab("") + ylab("") +
    ggtitle(ggtitle) +
    theme(legend.position = c(0.85,0.2),
          axis.text = element_text(size=14),
          legend.box.background = element_rect(fill = "white", linewidth=0.6, colour="gray"))

{ 
  if(x == "nb_dead_res") {
    
    p <- p + 
      geom_point(data=data %>% filter(estimate > plots_above[[1]]), aes(x=longitude,y=latitude,color=estimate), size=1.4) +
      geom_point(data=data %>% filter(estimate < plots_above[[2]]), aes(x=longitude,y=latitude,color=estimate), size=1.4) +
      geom_point(data=data %>% filter(estimate > plots_above[[3]]), aes(x=longitude,y=latitude,color=estimate), size=1.4) + 
      geom_point(data=data %>% filter(estimate < plots_above[[4]]), aes(x=longitude,y=latitude,color=estimate), size=1.4) + 
      scale_color_stepsn(name = legendtitle, colours = rainbow(8), limits=NULL, breaks=c(-20,-10,0,10,20,30,40)) 
    
      } else {
        
        p <-  p + 
          geom_point(data=data %>% filter(estimate>plots_above[[1]]), aes(x=longitude,y=latitude,color=estimate), size=1.4) +
          geom_point(data=data %>% filter(estimate>plots_above[[2]]), aes(x=longitude,y=latitude,color=estimate), size=1.4) +
          scale_color_gradientn(name = legendtitle, colours = unbalanced_rainbow, limits=NULL)
        } 
      } 
}
```

Probability of mortality $p$ during the survey period in each plot.

```{r ModelM62ProbMortalityPred, fig.height=8, fig.width=8}
p <- map_pred_nfi(x = "p", 
                  ggtitle = NULL, 
                  legendtitle = "Probability of tree\nmortality during\nthe inventory period",
                  plots_above = c(0.2,0.4),
                  unbalanced_rainbow = c("#00FF66", "#0066FF",rep("#CCFF00",2),rep("orange",3),rep("#FF0000",9)),
                  method_snp_set_code = method_snp_set_code_i)

ggsave(plot = p, filename = here("figs/ValidationNFI/ModelM62_PredictedMortalityProbabilityInventoryPeriod_Map.pdf"), height = 6, width = 6)

p
```


Annual probability $p_{annual}$ of tree mortality

```{r ModelM62AnnualProbMortalityPred, fig.height=8, fig.width=8}
p <- map_pred_nfi(x = "p_annual", 
                  ggtitle = NULL, 
                  legendtitle = "Annual probability\nof tree mortality",
                  plots_above = c(0.1,0.2),
                  unbalanced_rainbow = c("#00FF66", "#0066FF",rep("#CCFF00",2),rep("orange",3),rep("#FF0000",9)),
                  method_snp_set_code = method_snp_set_code_i)

ggsave(plot = p, filename = here("figs/ValidationNFI/ModelM62_PredictedAnnualMortalityProbability_Map.pdf"), height = 6, width = 6)

p
```

Annual probability $p_{annual,go}$ of tree mortality in each plot with $DBH$ and $C$ fixed to the mean and country fixed for Spain.

```{r ModelM62AnnualProbMortalityPredGO, fig.height=8, fig.width=8}
p <- map_pred_nfi(x = "p_annual_go", 
                  ggtitle = NULL, 
                  legendtitle = "Annual probability\nof tree mortality",
                  plots_above = c(0.065,0.08),
                  unbalanced_rainbow = c("#00FF66", "#0066FF",rep("#CCFF00",2),rep("orange",3),rep("#FF0000",5)),
                  method_snp_set_code = method_snp_set_code_i)

ggsave(plot = p, filename = here("figs/ValidationNFI/ModelM62_PredictedAnnualMortalityProbability_varyingGO_OtherParamsFixed_Map.pdf"), height = 6, width = 6)

p
```


$m_{pred}$ the predicted number of dead trees in each plot during the survey period.

```{r ModelM62PredicteNumberDeadTrees, fig.height=8, fig.width=8}
p <- map_pred_nfi(x = "nb_dead_pred", 
                  ggtitle = NULL, 
                  legendtitle = "Predicted number\nof dead trees",
                  unbalanced_rainbow = c("#00FF66", "#0066FF",rep("#CCFF00",2),rep("orange",3),rep("#FF0000",9)),
                  plots_above = c(8,14),
                  method_snp_set_code = method_snp_set_code_i)

ggsave(plot = p, filename = here("figs/ValidationNFI/ModelM62_PredictedNumberDeadTrees_Map.pdf"), height = 6, width = 6)

p
```

$r_{pred}$ the residuals (in each plot during the survey period): the observed number of dead trees $m_i$ minus the predicted number of dead trees $m_{pred}$.

```{r ModelM62PredictedResiduals, fig.height=8, fig.width=8}
p <- map_pred_nfi(x = "nb_dead_res", 
                  ggtitle = NULL, 
                  legendtitle = "Residuals",
                  unbalanced_rainbow = NULL,
                  plots_above = c(10,-10,20,-20),
                  method_snp_set_code = method_snp_set_code_i)

ggsave(plot = p, filename = here("figs/ValidationNFI/ModelM62_Residuals_Map.pdf"), height = 6, width = 6)

p
```

# Correlation coefficients 

We estimate correlations between genomic offset predictions and the observed annual mortality rates, after correction for confounding factors.

We estimate two types of correlations:

  - between genomic offset values and the observed annual mortality rates (after correction)
  
  - between the plot ranks based on the genomic offset predictions and the plot ranks based on the observed annual mortality rates (after correction).


```{r CompileMortalityModelM62WithPredictionsWithoutGO, message=F, warning=F, eval=T}
# Stan code
stancode = stan_model(here("scripts/StanModels/ValidationNFI_m6_2_with_predictions_withoutGO.stan"))
print(stancode)
```

```{r RunModelM62WithPredictionsWithoutGO, eval=F}
# Data in a list for Stan 
stanlist <- list(N = nrow(nfi_data),
                 nb_dead = nfi_data$nb_dead,
                 nb_tot=nfi_data$nb_tot,
                 C = (nfi_data$basal_area - mean(nfi_data$basal_area))/sd(nfi_data$basal_area),
                 DBH = (nfi_data$mean_DBH - mean(nfi_data$mean_DBH))/sd(nfi_data$mean_DBH),
                 nb_country=length(unique(nfi_data$country)),
                 country=as.numeric(as.factor(nfi_data$country)),
                 log_nb_years=log(nfi_data$nb_years))

       
# Running the Stan model
mod <- sampling(stancode, 
                data = stanlist, 
                iter = 2000, 
                chains = 4, 
                cores = 4,
                init=0, 
                save_warmup = FALSE,
                warmup = 1000, 
                verbose =T)

# Save the model
saveRDS(mod, file=here(paste0("outputs/ValidationNFI/StanModels/M62_WithoutGO.rds")))
```

```{r BuildCoeffTableModelM62WithPredictionsWithoutGO, eval=F}
mod <- readRDS(file=here("outputs/ValidationNFI/StanModels/M62_WithoutGO.rds"))

params_to_estimate <- "p_annual"

# Model coefficients
coefftab <- broom.mixed::tidyMCMC(mod,
                                  pars=params_to_estimate,
                                  droppars = NULL, 
                                  robust = FALSE, # give mean and standard deviation
                                  ess = T, # effective sample size estimates
                                  rhat = T, # Rhat estimates
                                  conf.int = T, 
                                  conf.level = 0.95 # include 95% credible intervals
                                 ) %>% 
  dplyr::rename(conf95_low=conf.low,
                conf95_high=conf.high,
                mean=estimate,
                std_deviation=std.error)
  
coefftab <- broom.mixed::tidyMCMC(mod,
                                  pars=params_to_estimate,
                                  droppars = NULL, 
                                  robust = TRUE, # give median and mean absolute deviation (= avg distance btw each data point and the mean)
                                  ess = F, 
                                  rhat = F, 
                                  conf.int = T, 
                                  conf.level = 0.80 # include 80% credible intervals
                                  ) %>% 
  dplyr::rename(conf80_low=conf.low,
                conf80_high=conf.high,
                median=estimate,
                mean_absolute_deviation=std.error) %>%
    inner_join(coefftab,by=c("term"))
  
coefftab_nfi <- nfi_data %>% 
  bind_cols(coefftab) %>% 
  mutate(res_mortality_median = annual_prop_dead - median) %>% 
  arrange(res_mortality_median) %>% 
  mutate(rank_res_mortality_median = 1:nrow(.)) %>% 
  mutate(res_mortality_mean = annual_prop_dead - mean) %>% 
  arrange(res_mortality_mean) %>% 
  mutate(rank_res_mortality_mean = 1:nrow(.))

# save the table
saveRDS(coefftab_nfi, file=here(paste0("outputs/ValidationNFI/M62_CoeffTable_WithoutGO.rds")))
```

```{r LoadCoeffTableModelM62WithPredictionsWithoutGO}
coefftab_nfi <- readRDS(file=here(paste0("outputs/ValidationNFI/M62_CoeffTable_WithoutGO.rds")))
```

Correlation between the mean and median estimates of the annual probability of mortality that can be explained by the confounding factors (i.e. competition, mean DBH, or country):

```{r CorrelationsMeanAndMedianModelM62WithPredictionsWithoutGO}
cor(coefftab_nfi$mean, coefftab_nfi$median)
```

Correlation between plot ranks of the observed annual mortality rates, after correcting for confounding factors, based on either the mean or the median:

```{r CorrelationsMeanAndMedianRanksModelM62WithPredictionsWithoutGO}
cor(coefftab_nfi$rank_res_mortality_mean,coefftab_nfi$rank_res_mortality_median)
```

Given the high correlations, we conclude that either the mean or the median can be used for calculating the correlations.

We calculate the correlations using the median estimates.

```{r CalculateCorrelationsModelM62WithPredictionsWithoutGO, message =F}
corrtab <- lapply(unique(df$method_snp_set_code), function(method_snp_set_i){
  
  sub_data <- df %>% 
    filter(method_snp_set_code == method_snp_set_i) %>% 
    #arrange(GO) %>% 
    #mutate(GO_rank = 1:nrow(.)) %>% 
    inner_join(coefftab_nfi, by="plotcode")
  
  tibble(pearson_correlation = cor.test(sub_data$res_mortality_mean, sub_data$GO, method="pearson")$estimate[[1]],
         pearson_pvalue = cor.test(sub_data$res_mortality_mean, sub_data$GO, method="pearson")$p.value,
         spearman_correlation = cor.test(sub_data$res_mortality_mean, sub_data$GO, method="spearman")$estimate[[1]],
         spearman_pvalue = cor.test(sub_data$res_mortality_mean, sub_data$GO, method="spearman")$p.value,
         method_snp_set_code = method_snp_set_i,
         method_snp_set_name = unique(sub_data$method_snp_set_name),
         snp_set_name = unique(sub_data$snp_set_name),
         snp_set_code = unique(sub_data$snp_set_code),
         method = unique(sub_data$method))
}) %>% 
  bind_rows() 

corrtab <- corrtab %>% 
  mutate(snp_set_name = factor(snp_set_name, levels = c(unique(corrtab$snp_set_name)[[6]],
                                                        unique(corrtab$snp_set_name)[[4]],
                                                        unique(corrtab$snp_set_name)[[5]],
                                                        unique(corrtab$snp_set_name)[[3]],
                                                        unique(corrtab$snp_set_name)[[1]],
                                                        unique(corrtab$snp_set_name)[[2]],
                                                        unique(corrtab$snp_set_name)[[7]],
                                                        unique(corrtab$snp_set_name)[[8]],
                                                        unique(corrtab$snp_set_name)[[9]],
                                                        unique(corrtab$snp_set_name)[[10]],
                                                        unique(corrtab$snp_set_name)[[11]],
                                                        unique(corrtab$snp_set_name)[[12]],
                                                        unique(corrtab$snp_set_name)[[13]])),
          method = factor(method, levels = c(unique(corrtab$method)[[1]],
                                             unique(corrtab$method)[[2]],
                                             unique(corrtab$method)[[3]],
                                             unique(corrtab$method)[[4]],
                                             unique(corrtab$method)[[5]],
                                             unique(corrtab$method)[[6]]))) 
```

**Comment:** When estimating the Spearman correlations, we get the following message:

`Message d'avis :
Dans cor.test.default(sub_data$res_mortality_mean, sub_data$GO, method = "spearman") :
  Impossible de calculer la p-value exacte avec des ex-aequos`
  
It indicates that the function cannot compute the exact p-value due to the presence of ties (i.e., identical values) in the data. This occurs because the Spearman correlation method is based on rank-ordering the data, and if two or more values are the same, it's unclear how to rank them uniquely. 
Spearman's correlation, being a non-parametric test, works by converting the data to ranks and then calculating the correlation between these ranks. When there are tied values (ex-aequos), the function adjusts for this, but the calculation of an exact p-value is more complex, so R resorts to an approximation.
In practical terms, the result is still valid, but the p-value is an approximation rather than the exact value. In our case, the number of ties is small (119 ties out of a total of 11,917 data points, where these 119 ties represent 47 unique values), this approximation should be close to the true value.


We plot the correlations.

```{r PlotCorrelationsModelM62WithPredictionsWithoutGO}
correlation_types <- c("pearson_correlation", "spearman_correlation")

lapply(correlation_types, function(coeff){
  
  p <- corrtab %>% 
    pivot_longer(values_to = "correlation_estimate", names_to = "correlation_type", cols = all_of(correlation_types)) %>% 
    mutate(correlation_pvalue = ifelse(correlation_type == "pearson_correlation", pearson_pvalue, spearman_pvalue)) %>% 
    dplyr::select(-pearson_pvalue,-spearman_pvalue) %>% 
    mutate(pvalue_signi = ifelse(correlation_pvalue < 0.05, "p-value < 0.05", "p-value ≥ 0.05")) %>% 
    filter(correlation_type == coeff) %>% 
    ggplot(aes(x = method,
               y = correlation_estimate,
               color = snp_set_name,
               shape = pvalue_signi)) +
    geom_rect(inherit.aes=FALSE, aes(xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf), color="transparent", fill="green", alpha=0.005) + 
    geom_hline(yintercept = 0,color="gray") +
    geom_point(position = position_dodge(width = .4), size=3) +
    {if(coeff=="pearson_correlation") ylab("Pearson correlation estimates")} + 
    {if(coeff=="spearman_correlation") ylab("Spearman correlation estimates (rank correlation)") } + 
    xlab("") +
    ylim(c(-0.2,0.2)) +
    scale_color_manual(values=colors_coeff) +
    scale_shape_manual(values = c(16,18)) + #, name="p-value of the correlations"
    theme_bw() +
    labs(color="",shape="") +
    theme(axis.text.x =  element_text(size=13),
          axis.text.y = element_text(size=13),
          axis.title.y = element_text(size=16),
          axis.title.x = element_text(size=1),
          legend.title=element_text(size=10), 
          #legend.box.background = element_rect(colour = "gray", linewidth=0.6),
          strip.text.x = element_text(size = 16),
          strip.background = element_blank(),
          #legend.position = c(0.63, 0.85),
          legend.text=element_text(size=9),
          panel.grid.minor.x=element_blank(),
          panel.grid.major.x=element_blank()) +
    guides(color=guide_legend(ncol=1),
           shape = guide_legend(override.aes = list(size =2 )))

p

# save in pdf and png
ggsave(p,file=here(paste0("figs/ValidationNFI/ModelM62_",coeff,"_CorrelationPlots.pdf")),
       device="pdf",
       height=7,
       width=12)

})
```

# GO predictions in Galicia

GO predictions in Galicia (especially those obtained with GF) show a strong boundary between two zones (see Figure 4b of the manuscript). How can we explain this pattern? 

```{r MapSurveyDate, fig.width=8,fig.height=8, results="hide"}
nfi_data %>% 
  mutate(year_second_survey = ifelse(second_survey > 1999, "After 2000",second_survey)) %>% 
           ggplot() + 
    geom_sf(data = world, fill="gray98") + 
    theme_bw() +
    scale_x_continuous(limits = c(-10, 12)) +
    scale_y_continuous(limits = c(35, 51)) + 
    geom_point( aes(x=longitude,
                    y=latitude,
                    color=factor(year_second_survey)),
               alpha=0.4, size=0.5) + 
    xlab("") + ylab("") +
    ggtitle("") +
    scale_color_manual(name = "Year of the second survey", 
                       values =  c("yellow2","darkorchid","#60F2E7","#C1E5AC")) +
    theme(legend.box.background = element_rect(colour = "gray80", linewidth=0.6)) +
   guides(colour = guide_legend(override.aes = list(size=2)))
```

In Galicia, GO predictions from the GF approach (see Figure 4b of the manuscript) are higher for plots surveyed for the second time in 1998 than plots surveyed for the second time in 1997.

We explore the climatic differences between the two sets of plots, ie plots surveyed in 1997 vs 1998. First, we extract the annual climatic values at the location of the Galician plots sampled in 1997 and 1998 and plot their values across the study period (ie between the first and second surveys).

```{r , fig.width=12,fig.height=8, results="hide", message =F}
# Extracting annual climatic values at the location of the NFI plots in Galicia surveyed for the second time inm 1997 or 1998
subclim <- clim %>% 
  left_join(nfi_data, by=c("plotcode","longitude","latitude")) %>% 
    dplyr::filter(second_survey %in% 1997:1998) %>% 
    dplyr::filter(year %in% min(first_survey):1998) %>% 
    dplyr::select(second_survey,year,all_of(clim_var))


# violin_plots <- lapply(clim_var, function(x) {
#   
# var_name <- extract_climatedt_metadata(var_clim = x) %>% 
#   mutate(var_legend= paste0(description, " (", label,"; ",unit_symbol,")")) %>% 
#   pull(var_legend)
# 
# subclim %>% 
#   mutate(year=as.factor(year)) %>% 
#   ggplot(aes_string(y=x,x="year")) + 
#   geom_jitter(shape=16,aes(colour=factor(second_survey)),position=position_jitter(0.2),size=2,alpha=0.3) +
#   xlab("") + 
#   scale_color_manual(name="Year of the second survey",
#                      values = c("yellow2","darkorchid")) +
#   ggtitle(var_name) +
#   ylab("mean climate under survey period - mean past climate") +
#   theme_bw()
# 
# })

violin_plots <- lapply(clim_var, function(x) {
  
var_name <- extract_climatedt_metadata(var_clim = x) %>% 
  mutate(var_legend= paste0(description, " (", label,"; ",unit_symbol,")")) %>% 
  pull(var_legend)

subclim %>% 
  mutate(year=as.factor(year),
         second_survey=as.factor(second_survey)) %>% 
  ggplot(aes_string(y=x,x="year",color="second_survey")) + 
  geom_point(shape=16,position=position_jitter(0.2),size=2,alpha=0.1) +
  geom_violin(alpha=0.3,fill="white", linewidth=0.7) + 
  xlab("") + ylab("") +
  ggtitle(var_name) +
  scale_color_manual(name="Year of the second survey",
                     values = c("yellow2","darkorchid")) +
  theme_bw() +
  guides(colour = guide_legend(override.aes = list(size=2)))

})

violin_plots 
```

From these plots, it seems that the plots sampled in 1997 have lower isothermality (`bio3`) and lower temperature seasonality (`bio4`).

We also compare the mean climatic values of the reference and survey periods between the plots surveyed in 1997 and 1998. For that, we calculate for each plot (and each climatic variable) the difference between its mean climate across the reference period (ie 1901-1950) and its mean climate across the survey period (between the first and second survey period).

```{r ViolinPlotsComparingRefandSurveyClimatesGalicianPlots, fig.height=6, fig.width=8, results="hide", message=F}
# We load the datasets with mean climatic values across the reference and survey periods
# Then we calculate the difference between the mean climates of the two periods for each plot
#  We then keep only plots surveyed (second survey) in 1997 or 1998 (Galician plots)
subnficlim <- readRDS(here("data/ClimaticData/NFIplots/NFIclimate.rds")) %>% 
  bind_rows() %>% 
  group_by(plotcode,longitude,latitude,elevation) %>% 
  summarise(across(tmn01:Eref, diff)) %>% 
  ungroup() %>% 
  left_join(nfi_data, by=c("plotcode","longitude","latitude")) %>%
  dplyr::filter(second_survey %in% 1997:1998)

bg_color <- "gray44" # background color

violin_plots <- lapply(clim_var, function(x) {
  
var_name <- extract_climatedt_metadata(var_clim = x) %>% 
  mutate(var_legend= paste0(description, " (", label,"; ",unit_symbol,")")) %>% 
  pull(var_legend)

subnficlim %>% 
  mutate(second_survey=as.factor(second_survey)) %>% 
  ggplot(aes_string(y=x,x="second_survey")) + 
  geom_violin(alpha=0.2, fill="gray76", color="darkgreen") + 
  geom_jitter(shape=16,position=position_jitter(0.2),size=2,alpha=0.4) +
  xlab("") + 
  ggtitle(var_name) +
  ylab("mean climate under survey period - mean past climate") +
  theme_bw()

})

violin_plots 
```

These plots show that plots surveyed in 1998 show higher increase in the mean annual temperature (`bio1`), lower increase in annual precipitation (`bio12`) and stronger decrease in temperature seasonality (`bio4`) than plots surveyed in 1997. As temperature seasonality is the most important variable to explain the allelic turnover in GF, it may explain the strong differences in GO predictions among plots surveyed in 1997 and 1998.

# Session information

```{r SessionInfo}
devtools::session_info()
```


