---
title: "Checking past and future climates at the location of the populations"
author: "Juliette Archambeau"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 10px
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7,fig.height = 5,cache=F)
options(width = 300)
library(knitr)      # CRAN v1.26
library(tidyverse)  # CRAN v1.3.0
library(readxl)     # CRAN v1.3.1
library(xtable)
library(reshape2)
library(kableExtra)
library(here)
library(raster)
library(magrittr)
library(cowplot)

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
source(here("scripts/functions/extract_climatedt_metadata.R")) # extracting meta data of the climatic variables in ClimateDT

# for the figures
bg_color <- "gray44"
```



```{bash, eval=F, echo=F}
# Bash commands that have been applied after the extraction

cd data/ClimaticData/ClimateDTRasters

# we remove some rasters to free space
find . -name "tmn*.tif" -exec rm {} \; # remove the 12 tmn*.tif files
find . -name "tmx*.tif" -exec rm {} \; # remove the 12 tmx*.tif files
find . -name "prc*.tif" -exec rm {} \; # remove the 12 prc*.tif files
find . -name "pet*.tif" -exec rm {} \; # remove the 12 pet*.tif files
find . -name "solRad*.tif" -exec rm {} \; # remove the 12 solRad*.tif files
find . -name "GDD*.tif" -exec rm {} \; # remove GDD0.tif and GDD18.tif
find . -name "*FFP.tif" -exec rm {} \; # remove eFFP.tif and FFP.tif
find . -name "PAS.tif"  -exec rm {} \; # remove PAS.tif

find . -name "MSP.tif" -execdir mv {} SP.tif \; # rename MSP.tif to SP.tif
```

# Goals

**Goal:** compare the values of the climatic variables at the location of the populations under current and future climates.


**Comment:** in all the plots below, the points corresponds to the 34 studied populations of maritime pine and the colors correspond to the main gene pool they belong to.


**Climatic variables of interest:**

```{r ListSelectedClimVariables}
# Selected climatic variables
# ===========================
clim_var <- readRDS(here("data/ClimaticData/NamesSelectedVariables.rds")) %>% 
  `[[`(3) # We keep the set 3 of climatic variables (the third list element)#

# clim_var <- readRDS(here("data/ClimaticData/NamesSelectedVariables.rds")) %>% 
#   lapply(function(x) x$variables) %>% 
#   unlist() %>% 
#   unique() %>% 
#   setdiff(c("bio15","bio9","Eref")) # no raster for bio15, bio9 and Eref

extract_climatedt_metadata(var_clim = clim_var$variables) %>% 
  dplyr::select(label,description,unit) %>% 
  set_colnames(str_to_title(colnames(.))) %>% 
  kable_mydf(font_size = 12)
```


# Past climates

## Reference periods

To account for the climates under which the populations have evolved, we could use the **reference period 1901-1950** or **1961-1990**. 

  
Until now, we have used the reference period 1901-1950 to capture as much as possible the climate under which the populations have evolved, i.e. before the human-induced climate change. Note that the human induced climate change may be minor before 1990 so we may use the pre-industrialization period 1961-1990 period, which may not be very different from the 1901-1950 period.

Another reason we used the reference period 1901-1950 is that our reference period has to be before 1983 because we have to use the gene-climate relationships estimated under this reference period to then predict the disruption of the gene-environment relationships under the climates between 1983 and 2014 at the location of the National Forest Inventory plots (one of the validation steps of the genomic offset method we do in the paper).

Note that the historic climatic layers for the 1901-1950 period may not be very accurate according to Maurizio Marchi, that's why he suggested to use the 1961-1990 period.

  
Below, we compare the mean values of the climatic variables of interest for the **reference period 1901-1950** and **1961-1990**.

```{r LoadPastClimateAndGenePoolInfo}
# Load population past climate
# The loaded file is a list with different reference periods, here 1901-1950 and 1961-1990
clim_past <- readRDS(here("data/ClimaticData/MaritimePinePops/PastClimatePopulationLocations.rds"))
 
# Load information about the main gene pool of each population (for the figures)
gps <- readRDS(here("data/GenomicData/MainGenePoolPopulations.rds")) %>% arrange(pop)

# Population coordinates as spatial points (for the extraction of future climates)
pop_coord <-  clim_past[[1]]$ref_means %>% 
  dplyr::select(longitude,latitude) %>% 
  SpatialPoints(proj4string=CRS("+proj=longlat +datum=WGS84 +no_defs"))
```



```{r FunctionGGScatterPlots}
make_ggscatter_plots <- function(df,x,gps,xlab,ylab){
  
var_name <- extract_climatedt_metadata(var_clim = x) %>% 
  mutate(var_legend= paste0(description, " (", label,"; ",unit_symbol,")")) %>% 
  pull(var_legend)
    
df %>% 
    left_join(gps, by="pop") %>% 
    mutate(main_gp_pop=factor(main_gp_pop,levels=c("Northern Africa",
                                                   "Corsica",
                                                   "Central Spain",
                                                   "French Atlantic",
                                                   "Iberian Atlantic",
                                                   "South-eastern Spain"))) %>% 
    
  ggplot(aes(x=x,y=y)) + 
  geom_point(aes(color=main_gp_pop),size=3) +
  scale_colour_manual(name="Main gene pool",
                      values = c("orangered3","gold2","darkorchid3","navyblue","turquoise2","green3")) +
  geom_abline(intercept = 0, slope = 1, color="gray60") +
  xlab(xlab) +
  ylab(ylab) +
  ggtitle(var_name) + 
  theme_bw() +
  theme(legend.position = "none",
        legend.title = element_text(size=16, color=bg_color),
        plot.title = element_text(size=16),
        legend.text = element_text(size = 16, color=bg_color),
        axis.text = element_text(size = 14, color=bg_color),
        axis.title = element_text(size=16, color=bg_color))
    
  
  }
```



```{r ScatterPlotsCompareRefPeriods, fig.height=11,fig.width=21, results="hide", warning=F}
# Generate the scatter plots
scatter_plots <- lapply(clim_var$variables, function(x){

  lapply(clim_past, function(y){
    
    y$ref_means %>% 
      dplyr::select(pop,all_of(x))
  }) %>% 
    bind_rows(.id="ref_period") %>% 
    pivot_wider(names_from = ref_period,values_from = x) %>% 
    set_colnames(c("pop","x","y"))  %>% 
    make_ggscatter_plots(x=x,
                       gps=gps,
                       xlab=paste0("Reference period ",clim_past[[1]]$range[[1]],"-",clim_past[[1]]$range[[2]]),
                       ylab=paste0("Reference period ",clim_past[[2]]$range[[1]],"-",clim_past[[2]]$range[[2]]))
})


p <- plot_grid(plotlist=scatter_plots)

# legend 
legend <- get_legend(scatter_plots[[1]] + theme(legend.position = c(0.5,0.5)) )

# merge plots + legend
plot_grid(p, legend, ncol = 2,rel_widths = c(1, 0.2))
```


<span style="color: orange;">**Interpretation:**</span>  The mean climatic values across the reference periods 1901-1950 and 1961-1990 are quite similar. We observe an increase in mean annual temperature for all populations across the 1961-1990 period compared to the 1901-1950 period. The patterns are more contrasted for the other climatic variables. For instance, populations from the French Atlantic gene pool experienced a decrease in isothermality while populations from the Iberian gene pool experienced an increase in isothermality.

## Point estimates vs raster extraction

The past climates can be generated in two ways:
  
  - point estimates provided by Maurizio Marchi at the location of the population. This corresponds to the climatic values we used in the graphs above.
  
  - extracted values from raster files. The pixel values in the raster file correspond to the mean values of the climatic variables across the period considered.
  
Below, we compare the point estimates climatic values (from csv) with extracted values from raster files for the period 1901-1950.


```{r ExtractClimVariables}
extract_clim_var <- function(x,gcm,period,ssp){
  
if(period=="1901-1950"){
  path <- paste0("data/ClimaticData/ClimateDTRasters/1km_1901-1950_Extent-Juliette/",x,".tif")} else {
  path <- paste0("data/ClimaticData/ClimateDTRasters/1km_",gcm,"_",period,"_",ssp,"_Extent-Juliette/",x,".tif")
  }
  
# For the variable Eref, there are some NAs for some populations
# So we start with a simple extraction (extraction of the cell value if the point falls within the cell)
# and if there are some NAs, we perform a bilinear extraction
  # bilinear interpolation = the returned values are interpolated from the values of the four nearest raster cells.
# And we replace the NAs by the extracted values obtained with the bilinear method
# if there are still NAs, we extract cell values using a buffer of 2km
  # If the distance between the sampling point and the center of a cell is less than or equal to the buffer, the cell is included. 
# We take the mean (fun=mean) of the cell values obtained with the extraction done with the buffer method
# And we replace the NAs by the values obtained with the buffer method (with a buffer of 2km)
# If there are still NAs, we do the same with a buffer of 3km
  
ext_simple <- raster::raster(here(path)) %>% raster::extract(pop_coord)
i <- is.na(ext_simple)

if(length(which(i==TRUE)) !=0){
  
ext_bilinear <- raster::raster(here(path)) %>% raster::extract(pop_coord,method="bilinear")
ext_simple[i] <- ext_bilinear[i]
i <- is.na(ext_simple)

if(length(which(i==TRUE)) !=0){
  
ext_buff <- raster::raster(here(path)) %>% raster::extract(pop_coord,buffer=2000,fun=mean)
ext_simple[i] <- ext_buff[i]
i <- is.na(ext_simple)

if(length(which(i==TRUE)) !=0){
  
ext_buff <- raster::raster(here(path)) %>% raster::extract(pop_coord,buffer=3000,fun=mean)
ext_simple[i] <- ext_buff[i]
i <- is.na(ext_simple)


}

}
  
}

return(ext_simple)

}
```


```{r ScatterPlotsPtEstimatesVsExtractedValues, fig.height=11,fig.width=21, results="hide", warning=F}
scatter_plots <- lapply(clim_var$variables, function(x) {
  
    
clim_past$ref_1901_1950$ref_means[,c("pop",x)] %>% 
    mutate(ext_var =extract_clim_var(x=x,period="1901-1950")) %>% 
    set_colnames(c("pop","x","y")) %>% 
    make_ggscatter_plots(x=x,
                       gps=gps,
                       xlab="Point estimates climatic values",
                       ylab="Extracted climatic values from raster")
  }) 

p <- plot_grid(plotlist=scatter_plots)

# legend 
legend <- get_legend(scatter_plots[[1]] + theme(legend.position = c(0.5,0.5)) )

# merge plots + legend
plot_grid(p, legend, ncol = 2,rel_widths = c(1, 0.2))
```

<span style="color: orange;">**Interpretation:**</span> Point estimate climatic values are **systematically higher** than the extracted values from rasters for the mean annual temperature and the isothermality. I found weird that the way climatic values are obtained (raster extraction vs point estimates) introduces a **systematic bias in the climatic values** (i.e. a bias towards systematic higher values for point estimate values). And this is annoying as it would mean that depending on which way of extracting the climatic data we decide to use, we will not have the same climatic shifts between past and future climates...


# Future climates

We aim to compare the predictions of five global climate models (GCMs) and quantify the differences between future and past climatic values.


## Using z-scores

To compare past and future climatic values, we calculate, for each GCM and each climatic variable, the mean $\mu_{ref}$ and standard deviation $\sigma_{ref}$ of the annual values of the climatic variable across the reference period (1901-1950 or 1961-1990). $\mu_{ref}$ and $\sigma_{ref}$ are then used to calculated a z-score, such as: *z-score* = $(x - \mu_{ref})/\sigma_{ref}$, where $x$ is the predicted mean value $X$ of the climatic variable of interest across the period 2041-2070.

On the graph below, the z-scores are on the y-axis and are calculated for four combinations of past and future climatic values:

  - either using the reference period 1901-1950 or 1961-1990. Note that in this section the past climatic values come from point estimates from csv files.
  
  - either using future climatic values extracted from raster files (corresponding to the mean predicted climatic values across the period 2041-2070 and the SSP3-7.0), or using future climatic values from point estimates from csv files (also for the period 2041-2070 and the SSP3-7.0). Note that those point estimates climatic values were calculated based on the population elevation data that we provided to Maurizio Marchi and not the elevation data from the ClimateDT DEEM, which may introduce some biaises.


```{r LoadFutureClimatePointEstimatesSentByMaurizio, warning=F}
# Point estimate future climatic values
fut_clim <- readRDS(here("data/ClimaticData/MaritimePinePops/FutureClimatePopulationLocationsPointEstimates.rds"))

# GCM names
gcm_names <- list.files(here("data/ClimaticData/ClimateDTRasters/")) %>%  str_sub(5,-34) %>% setdiff("")

# Extract population future climate for each GCM
clim_df <- lapply(clim_past, function(y){
  
fut_rast <- lapply(clim_var$variables, function(x){
  
lapply(gcm_names, function(gcm) extract_clim_var(x=x,gcm=gcm,period="2041-2070",ssp="ssp370")) %>% 
  setNames(gcm_names) %>% 
  as_tibble() %>% 
  mutate(pop=unique(clim_past[[1]]$ref_means$pop))

}) %>% 
  setNames(clim_var$variables) %>% 
  bind_rows(.id = "variable") %>% 
  pivot_longer(names_to = "gcm",values_to="mean_fut",cols=gcm_names) %>% 
  mutate(fut_method= "extracted from rasters")

fut_pe <- fut_clim %>% 
  bind_rows(.id="gcm") %>% 
  dplyr::select(pop,gcm,all_of(clim_var$variables)) %>% 
  pivot_longer(cols=all_of(clim_var$variables), names_to="variable",values_to="mean_fut") %>% 
  arrange(variable,pop) %>% 
  mutate(fut_method= "point estimates from csv")

y$zscores <- lapply(list(fut_rast,fut_pe), function(x){
  
x %>% left_join(y$ref_sdmeans, by=c("pop","variable")) %>% 
  dplyr::mutate(zscore=(mean_fut-mean_ref)/sd_ref) %>% 
  left_join(gps, by="pop") 
  })

return(y)
  
})
```


```{r CheckingNAs, eval=F}
# We check that there is no NAs:
lapply(clim_df, function(y) { lapply(y$zscores, function(z) {z %>%  sapply(function(x) sum(is.na(x))) })  })
```

```{r MakeViolinPlots, fig.height=12,fig.width=14, results="hide"}
# Violin plots

violin_plots <- lapply(clim_df, function(y){
  
lapply(y$zscores, function(z){


violin_plots <- lapply(gcm_names, function(x) {
  
z %>% 
  dplyr::filter(gcm %in% x) %>% 
  mutate(main_gp_pop=factor(main_gp_pop,levels=c("Northern Africa",
                                                 "Corsica",
                                                 "Central Spain",
                                                 "French Atlantic",
                                                 "Iberian Atlantic",
                                                 "South-eastern Spain"))) %>% 
  ggplot(aes(x=variable, y=zscore)) + 
  geom_hline(yintercept=0, color=bg_color) +
  geom_violin(alpha=0.2, fill="gray76", color=bg_color) + 
  geom_jitter(shape=16,aes(colour=main_gp_pop),position=position_jitter(0.2),size=1.7) +
  scale_colour_manual(values = c("orangered3",
                                                             "gold2",
                                                             "darkorchid3",
                                                              "navyblue",
                                                             "turquoise2",
                                                             "green3")) +
  xlab("") + 
  ylab("Z-scores") +
  labs(colour="Main gene pool") +
  theme_bw() +
  ggtitle(x) +
  theme(legend.position="none",
        legend.title = element_text(size=14, color=bg_color),
        legend.text = element_text(size = 14, color=bg_color),
        axis.text = element_text(size = 14, color=bg_color),
        axis.title = element_text(size=16, color=bg_color)) + 
  guides(colour = guide_legend(override.aes = list(size=4))) 
})


# legend 
violin_plots$legend <- get_legend(violin_plots[[1]] + theme(legend.position = c(0.5,0.5)) )

# make title
title <- ggdraw() + 
  draw_label(paste0("Future climate (", unique(z$fut_method),") vs reference period ",y$range[[1]],"-",y$range[[2]]),
             fontface = 'bold',
             x = 0,
             hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 7))

# merge plots
p <- plot_grid(plotlist=violin_plots)

# merge plots + title
plot_grid(title, p, ncol = 1,rel_heights = c(0.1, 1))
          
  
})

})

violin_plots
```

<span style="color: orange;">**Interpretation:**</span> Point estimates future climate values are **very different** from the distribution of past climatic values extracted from raster files (very high z-scores), which is not the case when we use future climatic values extracted from raster files, <span style="color: red;">**how can we explain that??**</span> Another thing that bothers me is that the mean annual temperature (`bio1`) and the mean temperature of the driest quarter (`bio9`) are predicted to decrease for most populations under some GCMs, I am a bit worried that there is something wrong somewhere..


## Raw differences between past and future climates

We aim to check whether the predicted decrease in mean annual temperature (`bio1`) and mean temperature of the driest quarter (`bio9`) for some populations can be due to the higher climatic values obtained with point estimates than with extracted values from rasters for the past climates (see first section)

We compare the differences between mean future climate and mean past climate using:

  - climatic values extracted from rasters for the future climates (as they appear more realistic according to the graph above)
  
  - past climatic values either from point estimates from csv files, or extracted values from rasters. 

```{r ViolinPlotsDiffFuturePastClimates,fig.height=12,fig.width=20, results="hide", warning=F}
clim_df$ref_1901_1950$zscores[[1]] <- lapply(clim_var$variables, function(x) {
  
    
clim_past$ref_1901_1950$ref_means[,"pop"] %>% 
    mutate(mean_ref_rast =extract_clim_var(x=x,period="1901-1950"),
           variable=x) 
  }) %>% 
  bind_rows() %>% 
  right_join(clim_df$ref_1901_1950$zscores[[1]],by = c("pop", "variable"))




clim_df$ref_1901_1950$zscores[[1]] %>% 
  pivot_longer(cols=c("mean_ref_rast","mean_ref"),names_to="mean_ref_meth",values_to = "mean_ref") %>% 
  group_split(mean_ref_meth) %>% 
  lapply(function(y){


violin_plots <- lapply(clim_var$variables, function(x) {
  
var_name <- extract_climatedt_metadata(var_clim = x) %>% 
  mutate(var_legend= paste0(description, " (", label,"; ",unit_symbol,")")) %>% 
  pull(var_legend)
  
y %>% 
    dplyr::filter(variable %in% x) %>% 
    dplyr::filter(!gcm %in% "Point-Estimates-GFDL-ESM4") %>% 
    mutate(diff=mean_fut-mean_ref) %>% 
    mutate(main_gp_pop=factor(main_gp_pop,levels=c("Northern Africa",
                                                 "Corsica",
                                                 "Central Spain",
                                                 "French Atlantic",
                                                 "Iberian Atlantic",
                                                 "South-eastern Spain"))) %>% 
  ggplot(aes(x=gcm, y=diff)) + 
  geom_hline(yintercept=0, color=bg_color) +
  geom_violin(alpha=0.2, fill="gray76", color=bg_color) + 
  geom_jitter(shape=16,aes(colour=main_gp_pop),position=position_jitter(0.2),size=2) +
  scale_colour_manual(values = c("orangered3",
                                                             "gold2",
                                                             "darkorchid3",
                                                              "navyblue",
                                                             "turquoise2",
                                                             "green3")) +
  xlab("") + 
  ylab("mean future climate - mean past climate") +
  labs(colour="Main gene pool") +
  theme_bw() +
  ggtitle(var_name) +
  theme(legend.position="none",
        legend.title = element_text(size=14, color=bg_color),
        legend.text = element_text(size = 14, color=bg_color),
        axis.text = element_text(size = 14, color=bg_color),
        axis.title = element_text(size=16, color=bg_color),
        ) + 
  guides(colour = guide_legend(override.aes = list(size=4))) 
})

p <- plot_grid(plotlist=violin_plots)

# legend 
legend <- get_legend(violin_plots[[6]] + theme(legend.position = c(0.5,0.5)) )

# merge plots + legend
p <- plot_grid(p, legend, ncol = 2,rel_widths = c(1, 0.2))


# make title
if(unique(y$mean_ref_meth=="mean_ref")){
  name_title <- "Past climatic values from point estimates (csv file)"
} else {
  name_title <- "Past climatic values extracted from raster files"
}

title <- ggdraw() + 
  draw_label(name_title,
             fontface = 'bold',
             x = 0,
             hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 7))

# merge plots + title
plot_grid(title, p, ncol = 1,rel_heights = c(0.1, 1))

})
```


<span style="color: orange;">**Interpretation:**</span> The differences between past and future climates do not show substantial differences depending on whether we use point estimates or extracted values from rasters for past climates. However, for mean annual temperature, we can see that there is a **clear trend towards more positive values** (higher temperatures in the future) when we use climatic values extracted from rasters for the past climate. 

<span style="color: red;">**FINAL COMMENT: So, we really have to think about which climatic values we use!!**</span>
