---
title: "Common garden data"
author: "Juliette Archambeau"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 11px
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
knitr::opts_chunk$set(cache=F, size="tiny")
options(width = 300)
library(knitr)      # CRAN v1.26
library(kableExtra) # CRAN v1.1.0
library(tidyverse)
library(here)

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
```

# Preparation file for ClimateDT

```{r GeneratePreExtractionFile}
coord <- read_csv(here("data/CommonGardenData/CommonGardenCoordinates.csv"), show_col_types = FALSE)
  
# As we do not have elevation data for the CG, we use the ClimateDT elevation data 
coord %>% 
  dplyr::select(site,latitude,longitude) %>% 
  arrange(site) %>% 
  mutate(elevation=case_when(site=="asturias" ~ 449,
                             site=="caceres" ~ 290,
                             site=="bordeaux" ~ 58,
                             site=="madrid" ~ 598,
                             site=="portugal" ~ 865)) %>% 
  write_csv(here("data/ClimaticData/CommonGardens/CommonGardensPreExtraction.csv"))
```

# Calculating mean climate for each site

Time period in each common garden:

  - Height measurement in **Asturias** (Spain) in November 2012 when the trees were 21 month-old: 2011 to 2012 (trees were planted in February 2011).
  
  - Height measurement in **Fundão** (Portugal) in October 2012 when the trees were 20 month-old: 2011 to 2012 (trees were planted in February 2011)
  
  - Height measurement in **Bordeaux** (France) in November 2018 when the trees were 85 month-old: 2012 to 2018 (trees were planted in October 2011).
  
  - Mortality and height measurements in **Madrid** (Spain) in December 2011 when the trees were 13 month-old: 2011 (trees were planted in November 2010).
  
  - Mortality and height measurements in **Caceres** (Spain) in December 2011 when the trees were 8 month-old: 2011 (trees were planted in April 2011).


```{r DefineCGmeasurementPeriods}
cg_periods <- list(asturias = 2011:2012,
                   portugal = 2011:2012,
                   bordeaux = 2012:2018,
                   caceres = 2011,
                   madrid = 2011)
```

```{r SetSelectedClimaticVariables}
# Set of climatic variables
# =========================
clim_var <- readRDS(here("data/ClimaticData/NamesSelectedVariables.rds"))
```


```{r CalculateMeanClimateForEachCG, warning=F, message=F}
# Function to calculate the average of the climatic variables at the location of the NFI plots
source(here("scripts/functions/calc_avg_clim_var.R"))

clim <- read_csv(here("data/ClimaticData/CommonGardens/ClimateDT_outputs.csv"), show_col_types = FALSE) %>% 
  dplyr::rename(cg=ID,
                longitude=Longitude,
                latitude=Latitude,
                elevation=Elevation,
                year=Year,
                SP=MSP) %>% 
  group_split(cg) %>% 
  purrr::map(\(x){
    
cg <- unique(x$cg)
cg_period <- cg_periods[[cg]]     
    
x %>% calc_avg_clim_var(ref_period = cg_period, id_spatial_points = "cg")

  }) %>% list_rbind() 


clim %>% saveRDS(here("data/ClimaticData/CommonGardens/ClimateCG.rds"))
```

